---
title: "Methods for Research and Policy Evaluation"
subtitle: "Session 5 : Regression discontinuity design"
author: "Paul Servais"
format: revealjs
---



## Last sessions

- Regressions
  - OLS
  - logistic
- Difference-in-differences

## Identification and exogeneity
- An OLS is a correlational analysis
\begin{equation}
Y_i = \alpha + \beta X_i + \delta D_i + \epsilon_i
\end{equation}
- Identifying a causal effect (directional) requires exogeneity and directionality
- $D \rightarrow Y$
- otherwise bias can arise

## Identification strategies
- RDD uses arbitrary decision rule as source of exogeneity
  - applies in settings where treatment is determined by whether an observed running variable $X$ crosses a known cutoff $c$.
- IV uses so-called instrument 
  - applies when treatment assignment is influenced by an exogeneous variable

# Regression discontinuity

## Sharp and fuzzy RDD
- Based on a running variable $X$
- Which influences probability $P_T(X\geq c)$ discontinuously
- Many exist out there because of legal thresholds
- Can be fuzzy or sharp
- Very close to the cutoff, individuals are treated/untreated
- But they also **look very similar**

## Intuition sharp RDD
- Age and drinking : deterministic
```{r}
#| echo: false

# Load library
library(ggplot2)

# Set seed
set.seed(123)

# Parameters
n <- 500
c <- 0  # cutoff

# Simulate running variable
X <- runif(n, -2, 2)

# Sharp treatment assignment (Heaviside)
D <- ifelse(X >= c, 1, 0)

# Add slight vertical jitter for visualization (so points don't overlap perfectly)
D_jitter <- D + rnorm(n, mean = 0, sd = 0.03)

# Dataframe
df <- data.frame(X, D, D_jitter)

# Create smooth grid for step function
X_grid <- seq(-2, 2, length.out = 1000)
P <- ifelse(X_grid >= c, 1, 0)
df_step <- data.frame(X_grid, P)

# Plot
ggplot() +
  # Cloud of points
  geom_point(data = df, aes(x = X, y = D_jitter), alpha = 0.4) +
  
  # Heaviside step function
  geom_step(data = df_step, aes(x = X_grid, y = P), size = 1) +
  
  # Cutoff line
  geom_vline(xintercept = c, linetype = "dashed") +
  
  labs(
    title = "Sharp RDD: Treatment Probability with Discontinuity",
    x = "Running Variable (X)",
    y = "Treatment (D)"
  ) +
  
  theme_minimal()
```

## Intuition fuzzy RDD
- Result at an entrance exam and admission rule (informal - yet not decisive threshold at 16/20). 
```{r}
#| echo: false

# Load library
library(ggplot2)

# Set seed
set.seed(123)

# Parameters
n <- 500
c <- 0  # cutoff
slope <- 5  # steepness of logistic curve
jump <- 0.3 # size of discontinuity at cutoff

# Simulate running variable
X <- runif(n, -2, 2)

# Logistic probability function
P <- 1 / (1 + exp(-slope * X))

# Add a discontinuity at cutoff
P[X >= c] <- P[X >= c] + jump
P <- pmin(P, 1)  # cap at 1

# Simulate treatment assignment
D <- rbinom(n, 1, P)

# Add vertical jitter for visualization
D_jitter <- D + rnorm(n, mean = 0, sd = 0.03)

# Dataframe for points
df <- data.frame(X, D_jitter)

# Smooth curve for theoretical probability
X_grid <- seq(-2, 2, length.out = 1000)
P_grid <- 1 / (1 + exp(-slope * X_grid))
P_grid[X_grid >= c] <- P_grid[X_grid >= c] + jump
P_grid <- pmin(P_grid, 1)
df_curve <- data.frame(X_grid, P_grid)

# Plot
ggplot() +
  # Cloud of points
  geom_point(data = df, aes(x = X, y = D_jitter), alpha = 0.4) +
  
  # Smooth fuzzy probability curve
  geom_line(data = df_curve, aes(x = X_grid, y = P_grid), color = "blue", size = 1.2) +
  
  # Cutoff line
  geom_vline(xintercept = c, linetype = "dashed") +
  
  labs(
    title = "Fuzzy RDD: Probability of Treatment with Discontinuity",
    x = "Running Variable (X)",
    y = "Pr(D = 1 | X)"
  ) +
  
  theme_minimal()
```

## Measuring the jump at discontinuity
```{r}
#| echo: false

library(tidyverse)
simulated_rdd_data_before_treatment <- 
  tibble(
    outcome = rnorm(200, mean = 1, sd = 1),
    running_variable = c(1:200),
    treatment = "before treatment"
  ) 

simulated_rdd_data_after_treatment <- 
  tibble(outcome = rnorm(200, mean = 3, sd = 1),
        running_variable = c(201:400),
        treatment = "after treatment"
  )

rbind(simulated_rdd_data_before_treatment, 
      simulated_rdd_data_after_treatment) |> 
  ggplot(aes(x = running_variable, y = outcome, color = treatment)) +
  geom_point(alpha = 0.5) +
  geom_smooth(formula = y ~ x, method = "loess") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_color_brewer(palette = "Set1")
```

## Example
- Scholarship and GPA
```{r}
#| echo: false
# Set seed for reproducibility
set.seed(123)

# Parameters
n <- 300           # number of students
cutoff <- 80       # scholarship cutoff
slope <- 0.05      # linear trend of GPA with score
jump <- 0.5        # effect of scholarship on GPA

# Simulate exam scores
X <- runif(n, 60, 100)

# Simulate GPA: linear trend + jump at cutoff + noise
Y <- 2 + slope * X + ifelse(X >= cutoff, jump, 0) + rnorm(n, 0, 0.2)

# Create dataframe
df <- data.frame(X, Y)
```

```{r}
#| echo: false

ggplot(df, aes(x = X, y = Y)) +
  geom_point(alpha = 0.6) +
  # Separate regression lines on each side of cutoff
  geom_smooth(data = subset(df, X < cutoff), aes(x = X, y = Y), method = "lm", color = "blue", se = FALSE) +
  geom_smooth(data = subset(df, X >= cutoff), aes(x = X, y = Y), method = "lm", color = "red", se = FALSE) +
  # Vertical line at cutoff
  geom_vline(xintercept = cutoff, linetype = "dashed") +
  labs(
    title = "RDD: Effect of Scholarship on GPA",
    x = "Entrance Exam Score",
    y = "GPA after 1 Year"
  ) +
  theme_minimal()
```

## Impact on the outcome
- This jump helps assess the impacts of the treatment on the outcome
\begin{equation}
\tau = \lim_{x \to c^+} \mathbb{E}[Y_i \mid X_i = x] - \lim_{x \to c^-} \mathbb{E}[Y_i \mid X_i = x]
\end{equation}
- $c^+ = c+h$ and $c^- = c-h$ with $h \rightarrow 0$
- Trade-off between
  - bias ($h \text{ big} \rightarrow \text{bias}$) 
  - statistical power ($h \text{ small} \rightarrow n \text{ small}$)

## Trade-off bias - variance
- Two bandwidths : 1 or 10
```{r}
ggplot(df, aes(x = X, y = Y)) +
  # Shaded bandwidth = 10 (lighter grey, wider)
  geom_rect(aes(xmin = cutoff - 10, xmax = cutoff + 10,
                ymin = -Inf, ymax = Inf),
            fill = "grey80", alpha = 0.4, inherit.aes = FALSE) +
  
  # Shaded bandwidth = 1 (darker grey, narrower)
  geom_rect(aes(xmin = cutoff - 1, xmax = cutoff + 1,
                ymin = -Inf, ymax = Inf),
            fill = "grey50", alpha = 0.5, inherit.aes = FALSE) +
  
  geom_point(alpha = 0.6) +
  
  # Separate regression lines
  geom_smooth(data = subset(df, X < cutoff),
              method = "lm", color = "blue", se = FALSE) +
  geom_smooth(data = subset(df, X >= cutoff),
              method = "lm", color = "red", se = FALSE) +
  
  geom_vline(xintercept = cutoff, linetype = "dashed") +
  
  labs(
    title = "RDD: Effect of Scholarship on GPA",
    x = "Entrance Exam Score",
    y = "GPA after 1 Year"
  ) +
  theme_minimal()
```
## LATE
- The data set will be restricted around the cutoff
- A local coefficient will be estimated
- Local Average Treatment Effect
- Can be parametric (you specify the relationship) or non parametric

## Example {.scrollable}

\begin{equation}
Y_{i\{h\}} = \alpha + \beta X_{i\{h\}} + \delta D_{i\{h\}} + \epsilon_i
\end{equation}

```{r}
#| echo: true

library(tidyverse)
library(MASS)
library(stargazer)
# Define the bandwidth
h <- 10  

# Subset data within ± the threshold
rd_data1<- df |> 
  filter(abs(X - 80) <= h)

rd_data2<- df |> 
  filter(abs(X - 80) <= 0.1*h)


# Create running variable centered at the threshold
rd_data1 <- rd_data1 |> 
  mutate(
    centered_X = X - 80,  # Centering at the threshold
    treatment = if_else(X >= 80, 1, 0) # Treatment indicator
  )

rd_data2 <- rd_data2 |> 
  mutate(
    centered_X = X - 80,  # Centering at the threshold
    treatment = if_else(X >= 80, 1, 0) # Treatment indicator
  )

# Estimate local linear regression with interaction
rd_model1 <- lm(Y~ treatment + centered_X, data = rd_data1)
rd_model2 <- lm(Y~ treatment + centered_X, data = rd_data2)

#Compare the models

stargazer(rd_model1, rd_model2, type = "text")
```

## Using built-in functions
```{r}
#| echo: true

library(rdrobust)
# Parameters

# Run robust RDD
rdd_result <- rdrobust(y = df$Y, x = df$X, c = cutoff, h = 10) # h = points allows to specify the number of points

# Print summary
summary(rdd_result)
```

## Continuity at the cutoff {.scrollable}
- Absent of treatment, $Y$ would have followed a continuous trend at the cutoff. 
- Typical tests
  - **Covariate continuity tests** : placebos with other covariates
  - **McCrary density test**: test whether the density of the running variable is continuous at the cutoff, which helps detect manipulation or sorting around the threshold. 
  - **Placebos** with other cutoffs
  - **Varying bandwidths**: verify that results are robust to different choices of bandwidth.
  - **Varying polynomial orders**: check that results are not driven by a specific functional form, by changing the order of the estimation in the argument of the function `p = `.


## Covariate continuity test

- In a valid **RDD**, units just below and above the cutoff should be **similar**
- This means **pre-treatment covariates** must be **continuous at the threshold**
  
  - The design may be **invalid**
  - There could be **manipulation or sorting**


## What do we test?
- We check whether:
- Age, income, education, etc.
- change **smoothly** at the cutoff

## Example

```{r}
#| echo: true
#| eval: false
library(tidyverse)

set.seed(123)

dat_good <- tibble(
  x = runif(1000, 0, 100),
  D = ifelse(x >= 50, 1, 0),
  age = 30 + 0.1*x + rnorm(1000, 0, 2)  # smooth
)

ggplot(dat_good, aes(x, age, color = factor(D))) +
  geom_point(alpha = 0.3) +
  geom_vline(xintercept = 50, linetype = 2) +
  stat_smooth(method = "loess", se = FALSE) +
  labs(title = "No jump at cutoff",
       x = "Running variable",
       y = "Covariate (Age)")
```

## Example

```{r}
#| echo: false
library(tidyverse)

set.seed(123)

dat_good <- tibble(
  x = runif(1000, 0, 100),
  D = ifelse(x >= 50, 1, 0),
  age = 30 + 0.1*x + rnorm(1000, 0, 2)  # smooth
)

ggplot(dat_good, aes(x, age, color = factor(D))) +
  geom_point(alpha = 0.3) +
  geom_vline(xintercept = 50, linetype = 2) +
  stat_smooth(method = "loess", se = FALSE) +
  labs(title = "No jump at cutoff",
       x = "Running variable",
       y = "Covariate (Age)")
```


## McCrary density test {.scrollable}
- You expect similar density on both sides of cutoff
- Imagine some students manage to convince or corrupt their profs
- Just call `dens_test <- rddensity(X = df$X, c = cutoff)` from `rddensity` package
- NULL = continuity at the cutoff
- reject if p-value < 0.05
```{r}
#install.packages("rddensity")
# Load packages

library(rddensity)
library(ggplot2)

set.seed(123)

# Parameters
n <- 500
cutoff <- 80

# Simulate running variable X (exam score)
# Normal distribution around 75–85
X <- rnorm(n, mean = 78, sd = 5)

# Introduce manipulation: add "bunching" just above the cutoff
# Students who scored 78–79 get bumped to 80
manipulated <- X > 77 & X < cutoff
X[manipulated] <- cutoff

# Simulate outcome variable Y (GPA), continuous around 3.0
Y <- 2.5 + 0.03 * X + rnorm(n, 0, 0.2)

df <- data.frame(X, Y)

# Run McCrary density test
dens_test <- rddensity(X = df$X, c = cutoff)

# Summary
summary(dens_test)

# Plot the density
rdplotdensity(dens_test, X = df$X)

```
## Biased 
- Sorting around the threshold
```{r}
ggplot(df, aes(x = X, y = Y)) +
  # Shaded bandwidth = 10 (lighter grey, wider)
  geom_rect(aes(xmin = cutoff - 10, xmax = cutoff + 10,
                ymin = -Inf, ymax = Inf),
            fill = "grey80", alpha = 0.4, inherit.aes = FALSE) +
  
  # Shaded bandwidth = 1 (darker grey, narrower)
  geom_rect(aes(xmin = cutoff - 1, xmax = cutoff + 1,
                ymin = -Inf, ymax = Inf),
            fill = "grey50", alpha = 0.5, inherit.aes = FALSE) +
  
  geom_point(alpha = 0.6) +
  
  # Separate regression lines
  geom_smooth(data = subset(df, X < cutoff),
              method = "lm", color = "blue", se = FALSE) +
  geom_smooth(data = subset(df, X >= cutoff),
              method = "lm", color = "red", se = FALSE) +
  
  geom_vline(xintercept = cutoff, linetype = "dashed") +
  
  labs(
    title = "RDD: Effect of Scholarship on GPA",
    x = "Entrance Exam Score",
    y = "GPA after 1 Year"
  ) +
  theme_minimal()
```

## Polynomial orders {.scrollable}
- Why increase the degrees of your polynoms?
```{r}
# simultate nonlinearity
dat <- tibble(
    x = rnorm(1000, 100, 50)
  ) %>% 
  mutate(
    x = case_when(x < 0 ~ 0, TRUE ~ x),
    D = case_when(x > 140 ~ 1, TRUE ~ 0),
    x2 = x*x,
    x3 = x*x*x,
    y3 = 10000 + 0 * D - 100 * x + x2 + rnorm(1000, 0, 1000)
  ) %>% 
  filter(x < 280)


p1 <- ggplot(aes(x, y3, colour = factor(D)), data = dat) +
  geom_point(alpha = 0.2) +
  geom_vline(xintercept = 140, colour = "grey", linetype = 2) +
  stat_smooth(method = "lm", se = F) +
  labs(x = "Test score (X)", y = "Potential Outcome (Y)")

p2 <- ggplot(aes(x, y3, colour = factor(D)), data = dat) +
  geom_point(alpha = 0.2) +
  geom_vline(xintercept = 140, colour = "grey", linetype = 2) +
  stat_smooth(method = "loess", se = F) +
  labs(x = "Test score (X)", y = "Potential Outcome (Y)")

p1
p2
```

## Visualising

```{r}
library(rdrobust)
set.seed(123)

# Parameters
n <- 500
cutoff <- 80

# Running variable
X <- runif(n, 60, 100)

# Nonlinear baseline + treatment effect + noise
Y <- 1 + 0.04 * X - 0.0003 * X^2 +   # nonlinear trend
     0.8 * (X >= cutoff) +           # treatment effect
     rnorm(n, 0, 0.3)

df <- data.frame(X, Y)

library(dplyr)
library(ggplot2)

# Create dataset with polynomial labels
df_list <- list(
  df %>% mutate(p = "p = 1"),
  df %>% mutate(p = "p = 2"),
  df %>% mutate(p = "p = 3")
)

df_all <- bind_rows(df_list)

ggplot(df_all, aes(x = X, y = Y)) +
  geom_point(alpha = 0.3) +
  
  # p = 1
  geom_smooth(
    data = subset(df_all, p == "p = 1" & X < cutoff),
    method = "lm", formula = y ~ x,
    se = FALSE, color = "blue"
  ) +
  geom_smooth(
    data = subset(df_all, p == "p = 1" & X >= cutoff),
    method = "lm", formula = y ~ x,
    se = FALSE, color = "blue", linetype = "dashed"
  ) +
  
  # p = 2
  geom_smooth(
    data = subset(df_all, p == "p = 2" & X < cutoff),
    method = "lm", formula = y ~ poly(x, 2),
    se = FALSE, color = "red"
  ) +
  geom_smooth(
    data = subset(df_all, p == "p = 2" & X >= cutoff),
    method = "lm", formula = y ~ poly(x, 2),
    se = FALSE, color = "red", linetype = "dashed"
  ) +
  
  # p = 3
  geom_smooth(
    data = subset(df_all, p == "p = 3" & X < cutoff),
    method = "lm", formula = y ~ poly(x, 3),
    se = FALSE, color = "green"
  ) +
  geom_smooth(
    data = subset(df_all, p == "p = 3" & X >= cutoff),
    method = "lm", formula = y ~ poly(x, 3),
    se = FALSE, color = "green", linetype = "dashed"
  ) +
  
  geom_vline(xintercept = cutoff, linetype = "dashed") +
  
  facet_wrap(~p) +
  
  labs(
    title = "RDD Fits by Polynomial Order",
    subtitle = "Solid = Left of cutoff, Dashed = Right",
    x = "Running Variable",
    y = "Outcome"
  ) +
  theme_minimal()

y <- df$Y
x <- df$X

```
## Using rdrobust
```{r}
#| echo: true
# Different polynomial orders
rdd_p1 <- rdrobust(y, x, c = cutoff, p = 1)
rdd_p2 <- rdrobust(y, x, c = cutoff, p = 2)
rdd_p3 <- rdrobust(y, x, c = cutoff, p = 3)

results <- data.frame(
  p = c(1, 2, 3),
  tau = c(rdd_p1$Estimate[1], rdd_p2$Estimate[1], rdd_p3$Estimate[1]),
  se  = c(rdd_p1$se[1], rdd_p2$se[1], rdd_p3$se[1]),
  pval = c(rdd_p1$pv[1], rdd_p2$pv[1], rdd_p3$pv[1])
)
results

```

# Exercices
