---
title: Session 2 — Relationships between variables
toc: true
toc-depth: 3
editor: 
  markdown: 
    wrap: 72
---

# Linear and non-linear regressions

## Motivation

## Types of relationships

Of course, in practice we are often interested in studying relationships between different types of variables, and the choice of statistical method depends on the measurement scale of both the independent and dependent variables. When both the independent and dependent variables are categorical (for example, gender and voting preference, coded at the individual level as 0 and 1), we typically use crosstabulation and chi-square tests to assess whether the distribution of one variable differs across the categories of the other. If the independent variable is categorical but the dependent variable is continuous (for example, comparing average incomes across two or more groups), we use mean comparison methods such as the t-test to determine whether group means differ. When the independent variable is continuous and the dependent variable is categorical (for example, predicting whether someone purchases a product based on age or income), logistic regression (often expressed in terms of logits) allows us to model the probability of category membership as a function of the continuous predictors. Finally, when both variables are continuous, we can assess their association with correlation and explore predictive relationships with regression analysis. Understanding these distinctions helps you choose appropriate analytical tools that match the nature of your data and research questions.

Table: Recommended types of analysis based on variable types

|                         | **DV: Categorical**              | **DV: Interval/Continuous**          |
|-------------------------|-----------------------------------|----------------------------------------|
| **IV: Categorical**                | **Crosstabs / χ² Test**          | **Compare Means / t-Test**             |
| **IV: Interval/Continuous**        | **Logistic Regression / Logits** | **Correlation / Scatter / Regression**  |


In the previous session, we examined how to conduct t-tests and perform correlation and linear regression analyses—two cases in which the dependent variable is continuous. However, we did not discuss the conditions under which linear regression provides a good approximation of the relationship between two variables, nor did we consider situations in which the dependent variable is categorical.

This session therefore begins by reviewing the main assumptions underlying linear regression. It then introduces alternative models, such as logistic regression, which allow us to assess the impact of continuous independent variables on categorical dependent variables.

## Linear regressions : hypotheses
We have seen previously that a linear model can be written as : 

\begin{equation}
Y_i = \beta_0+ \beta_1 X_{i1} + \beta_2 X_{i2} + \beta_3 X_{i3} + ... + \beta_k X_{ik} + \epsilon_i
\end{equation}

where the $\beta_i$ coefficients are all constant - which is why it is a linear link between variable $Y$ and $X$.

## Assumptions
For this relation to hold, the following hypotheses must be true :

* $\mathbb{E} (\epsilon_i|X_i) = 0$. That basically means that the expectation of the error should be (there is no systematic error in your estimation), once you know $X_i$. This is called exogeneity : the error comes from elsewhere, and is not related to your variables. 
* $\mathbb{E} (\epsilon_i^2) = \sigma^2$. That means that the variance of the error is constant accross all values. This condition is called 'homoskedasticity'. The case where  $\mathbb{E} (\epsilon_i^2) \ne \sigma^2$ is called heteroskedasticity.
* $\mathbb{E} (\epsilon_i \epsilon_j)= 0$ or $\mathbb{C}ov(\epsilon_i, \epsilon_j) = 0$, which is called the linear independence of the errors assumption.
* $X_{ij}$ is non random $\forall  i = 1,...,n ; j = 1, ..., k$ (X is not a random variable, contrary to the error and the dependent variable).
* $\text{rank}((X_{ij})) = k$. This is the mathematical way to say that the different independent variables are indeed independent from one another. That means that there is no way to obtain any $X_j$ using sums and combinations of all the other $X_k$ (where $k \ne j$). In the econometric jargon, violating that assumption would lead to a model with so-called *multicollinearity* between predictors.

## Visualising the different assumptions

You may very legitimately feel puzzled by these mathematical expressions, as their meaning is not always immediately intuitive. This is why plotting graphs and working through concrete examples can help clarify what these assumptions really mean. Imagine that you are studying a physical phenomenon.

From physics, you know that the weight of an object is given by the formula:

\begin{equation}
W = m \times g,
\end{equation}

where $m$ is the mass of the object and $g$ is the acceleration due to gravity. On Earth, $g \approx 9.81 \, \frac{m}{s^2}$.

Now suppose you want to conduct a small experiment to verify this physical law using real data. For example, you could ask 100 students to each take an object with a mass between 0.01 kg and 10 kg and measure its weight using a dynamometer (an instrument that measures force in Newtons using a spring).

In theory, each student should obtain a measurement equal to:

\begin{equation}
W_i = m_i \times 9.81.
\end{equation}

In practice, however, the measurements will not be perfectly equal to this theoretical prediction. Measurement accuracy may vary depending on the student’s handling of the instrument, the precision of the dynamometer, calibration issues, or other small sources of error. All these sources of errors can lead to a total error, $\epsilon_i$, that should not be related to your prediction, in a systematic way.

For example, a student measuring a mass of 0.1 kg might record a weight of 0.96 N instead of 0.981 N, while another student measuring a mass of 10 kg might record 102 N instead of 98.1 N.

As a result, each observed weight $W_i$ will differ slightly from the theoretical prediction. We can write this as:

\begin{equation}
W_i = m_i \times 9.81 + \varepsilon_i,
\end{equation}

where $\varepsilon_i$ represents the error term—that is, the difference between the observed value and the true physical relationship. You could simulate such sample in R by plotting the following graph, where I assume that the errors have a normal distribution with constant standard deviation $\sigma$ (meaning that the values of your errors may vary). This would be the perfect example of a linear regression model.

```{r}
#| echo: true
#| eval: true
library(tidyverse)
library(tidyverse)

m <- sample(1:100, 100, replace = FALSE) / 10
W <- 9.81 * m + rnorm(100, 0, 15)

dat <- data.frame(m, W)
lm_model_biased <- lm(W ~ m, data = dat)
summary(lm_model_biased)

dat %>% 
  ggplot(aes(x = m, y = W)) + 
  geom_point() +
  geom_smooth(method = "lm",
              linetype = "dashed",
              color = "black",
              se = TRUE,
              linewidth = 1,
              alpha = 0.4)+
  theme_minimal()
```

The exogeneity condition means that these errors are not systematically related to the mass. In other words, the measurement errors are purely random and do not depend on whether the mass is small or large:

\begin{equation}
E(\varepsilon_i \mid m_i) = 0.
\end{equation}

This assumption ensures that, on average, the errors cancel out, allowing us to recover the true relationship between mass and weight using linear regression.

### When the exogeneity assumption is violated: biased measurement for large masses

The exogeneity assumption requires that the error term is not systematically related to the explanatory variable:

\begin{equation}
E(\varepsilon_i \mid m_i) = 0.
\end{equation}

This means that measurement errors must be purely random and must not depend on the mass being measured.

However, this assumption may be violated if the measurement instrument is imperfect in a systematic way. Suppose, for example, that the dynamometer is poorly calibrated for higher masses, and that the error of the calibrator is a function of $m^3$. This could happen if the spring inside the dynamometer becomes slightly stiffer when stretched too much, causing the instrument to overestimate the true weight for heavier objects.

In this case, the measurement error would no longer be random. Instead, it would increase with mass. Heavier objects would tend to have larger positive errors, while lighter objects would have smaller errors.

```{r}
#| echo: true
#| eval: true
library(tidyverse)
library(tidyverse)

m <- sample(1:100, 100, replace = FALSE) / 10
W <- 9.81 * m + rnorm(100, m**3, 15)

dat <- data.frame(m, W)
lm_model_biased <- lm(W ~ m, data = dat)
summary(lm_model_biased)

dat %>% 
  ggplot(aes(x = m, y = W)) + 
  geom_point() +
  geom_smooth(method = "lm",
              linetype = "dashed",
              color = "black",
              se = TRUE,
              linewidth = 1,
              alpha = 0.4)+
  theme_minimal()
```

We can still write the observed relationship as:

\begin{equation}
W_i = m_i \times 9.81 + \varepsilon_i,
\end{equation}

but now the error term depends on the mass:

\begin{equation}
E(\varepsilon_i \mid m_i) \neq 0.
\end{equation}

For example, the true physical relationship may be:

\begin{equation}
W_i^{true} = m_i \times 9.81,
\end{equation}

but the dynamometer introduces a systematic bias such as:

\begin{equation}
\varepsilon_i = 0.05 \times m_i + u_i,
\end{equation}

where \(u_i\) is a purely random error term satisfying:

\begin{equation}
E(u_i \mid m_i) = 0.
\end{equation}

In this situation, larger masses will tend to have larger errors on average. This creates a correlation between the error term and the explanatory variable:

\begin{equation}
\text{Cov}(m_i, \varepsilon_i) \neq 0.
\end{equation}

This violates the exogeneity assumption.

As a result, linear regression will no longer recover the true physical relationship. Instead of estimating the true coefficient of 9.81, it will estimate a biased coefficient that incorporates both the true physical effect and the measurement bias.

This example illustrates the key idea behind endogeneity: when the error term is correlated with the explanatory variable, the regression estimates become biased and do not reflect the true causal relationship. The error being endogenous basically means that your error term is actually caused by your independent variable. The error being exogeneous means that it is not caused by your independent variables.


## Homoskedasticity and heteroskedasticity

A second important assumption when conducting a linear regression is that the variance of the error term should be constant. This assumption is called **homoskedasticity**. It means that the variability of the error term should not depend on the value of the independent variable.

Formally, this assumption can be written as:

\begin{equation}
\text{Var}(\varepsilon_i \mid m_i) = \sigma^2
\end{equation}

for all values of $m_i$.

Returning to our example, suppose that your dynamometer provides very precise measurements for small masses, but becomes less reliable for larger masses due to poor calibration or mechanical limitations. In this case, measurements for small masses will be tightly clustered around the true value, while measurements for larger masses will be more dispersed.

This means that although the average relationship between mass and weight may still be correct, the variability of the errors increases with mass.

We can represent this situation as:

\begin{equation}
W_i = m_i \times 9.81 + \varepsilon_i
\end{equation}

but now the variance of the error depends on the mass:

\begin{equation}
\text{Var}(\varepsilon_i \mid m_i) = \sigma_i^2
\end{equation}

where $\sigma_i^2$ increases with $m_i$.

This situation is called **heteroskedasticity**.

Graphically, this results in a "fan-shaped" pattern: observations are tightly grouped for small masses and become increasingly spread out as mass increases.

Importantly, unlike violations of exogeneity, heteroskedasticity does not bias the estimated regression coefficients. However, it does affect the precision of the estimates and leads to incorrect standard errors, confidence intervals, and hypothesis tests if not properly accounted for. A solution to these problems is to run "robust linear models", through `rlm`. These account for the heteroskedasticity and compute better versions of the standard errors (and associated hypotheses). A good diagnostics for heteroskedasticity is also to plot `residuals` against `fitted.values` and to see if you can detect clear patterns in the data. This is examplified in the examples below.

```{r}
#| echo: true
#| eval: true
library(tidyverse)
library(tidyverse)

m <- sample(1:100, 100, replace = FALSE) / 10
W <- 9.81 * m + rnorm(100,0, 2*m)

dat <- data.frame(m, W)
lm_model_biased <- lm(W ~ m, data = dat)
summary(lm_model_biased)

dat %>% 
  ggplot(aes(x = m, y = W)) + 
  geom_point() +
  geom_smooth(method = "lm",
              linetype = "dashed",
              color = "black",
              se = TRUE,
              linewidth = 1,
              alpha = 0.4)+
  theme_minimal()
```


```{r}
library(performance)
library(MASS)
library(stargazer)

x <- sample(1:1000, 1000, replace = "TRUE")

y <- 10*x + rnorm(1000, 0, 10)
y <- data.frame(y, x)
model1 <- lm(y ~ x, data = y)

y <- 10*x + rnorm(1000, 0, 10*(x))
y <- data.frame(y, x)
model2 <- lm(y ~ x, data = y)
y %>% 
  ggplot(aes(x = x, y = y)) + geom_point()+
  geom_smooth(method = "lm", linetype = "dashed", color = "black", se = TRUE, linewidth = 1, alpha = 0.4)+
  theme_minimal()

mean(model2$residuals)
plot(model2$fitted.values, model2$residuals)
check_heteroscedasticity(model2) # Run a Breush-Pagan test

model2 <- lm(y ~ x, data = y)
# Fit robust regression
robust_model <- MASS::rlm(y ~ x, data = y)

stargazer(model1, model2, robust_model, align = TRUE, 
          type = "text")
```


## Multicollinearity 
The last typical problem that can arise in linear regressions is *multicollinearity*. This occurs when two explanatory variables are highly correlated with each other.

Returning to our weight-measurement example, suppose that in addition to mass, you would like to verify whether weight is influenced by the volume of the objects. To do this, you might ask all students to also compute the volume of their objects and include volume as a second explanatory variable in the regression.

Your regression model would then become:

\begin{equation}
W_i = \alpha m_i + \beta V_i + \varepsilon_i,
\end{equation}

where $m_i$ is the mass, $V_i$ is the volume, and $\varepsilon_i$ is the error term. Your goal would be to estimate the coefficients $\alpha$ and $\beta$.

However, suppose that all the objects are made of the same material, such as copper. In physics, mass and volume are mechanically related through the density:

\begin{equation}
m_i = \rho_c V_i,
\end{equation}

where $\rho_c$ is the density of copper (in $kg/m^3$). In practice, students may make small measurement errors when computing volumes, so the observed relationship becomes:

\begin{equation}
m_i = \rho_c V_i + \mu_i,
\end{equation}

where $\mu_i$ represents measurement error.

This means that mass and volume are almost perfectly correlated. As a result, the regression has difficulty distinguishing the individual effect of mass from the effect of volume, because both variables contain nearly the same information.

This situation is called *multicollinearity*.

Importantly, multicollinearity does not bias the estimated coefficients. However, it makes them much less precise: their standard errors increase, and the estimates may become unstable or highly sensitive to small changes in the data. As a result, it becomes difficult to determine which variable truly affects the outcome.

In our example, since weight is physically determined by mass, adding volume does not provide meaningful additional information, but it does make the estimation less precise because of the strong correlation between mass and volume. Multicollinearity makes it difficult to separate the individual effects of correlated predictors, because changing one predictor almost mechanically implies changing the other.

You can test for multicollinearity in R using the function `check_collinearity()`, which indicates whether your explanatory variables are highly correlated. This function typically reports the Variance Inflation Factor (VIF), a measure of how much the variance of a coefficient estimate is inflated due to multicollinearity. Large VIF values indicate problematic levels of collinearity between predictors.

When multicollinearity is present, one possible solution is to reduce the number of explanatory variables. This can be done using advanced data reduction techniques, such as Principal Component Analysis (PCA), which combines correlated variables into a smaller number of independent components.

However, an even more important approach is to rely on theory to guide variable selection.

In our example, mass and volume are highly correlated because all objects are made of copper, which has a constant density. From physics, we know that weight depends on mass, not on volume directly. Volume only affects weight indirectly through its relationship with mass.

Therefore, including both mass and volume in the regression does not provide meaningful additional information. Instead, it introduces multicollinearity and reduces the precision of the estimates. Based on theoretical considerations, the appropriate solution is to exclude volume from the regression and keep mass as the relevant explanatory variable.

This reasoning also applies to social science data. When two variables measure closely related concepts, or when one variable mechanically determines another, including both in a regression may create multicollinearity without improving the model. In such cases, theory and substantive knowledge should guide the choice of which variables to include.

```{r}
#| echo: true
library(performance)

set.seed(123)

m <- sample(1:100, 100, replace = FALSE) / 10

# density of copper ≈ 8920 kg/m^3
v <- m/8920 + rnorm(100, 0, 0.00001)  # small measurement error

y <- 9.81*m + rnorm(100, 0, 10)

model1 <- lm(y ~ m)
model2 <- lm(y ~ m + v)

summary(model1)
summary(model2)

check_collinearity(model2)
```

## Models with interactions

Another thing we didn't cover in the last session was interactions. You may now wonder how heterogeneous your results are, and how your variables interact. In our example with the far-right vote, perhaps the effect of income on the far-right vote is conditional on another variable, that is for instance the level of education ; and vice-versa. In other words, the results might be heterogeneous : a rich commune with less educated people can potentially vote differently than a rich commune with many higher educated persons. This is what is meant by checking for interactions between independent variables. For instance, in the case of a regression model with three variables, instead of checking :

\begin{equation}
Y_i = \beta_1 X_{1i} + \beta_2 X_{2i} + \beta_3 X_{3i},
\end{equation}

you try :
\begin{equation}
Y_i = \beta_1 X_{1i} + \beta_2 X_{2i} + \beta_3 X_{3i} + \beta_4 X_{1i}*X_{3i}
\end{equation}

Which means that any increase $\Delta X_{1i}$ in the variable $X_{1i}$ no longer has an impact $\beta_1 \Delta X_{1i}$ on $Y_i$, but an impact $\Delta X_{1i} (\beta_1 + \beta_3X_{3i})$, which therefore still depends on the variable $X_{3i}$. Coming back to our example, we want to assess how income and education in a commune evolve and influence far-right voting. You only need to specify `*` in your regression to obtain the interaction, and we generally use `ggpredict()` to plot the predicted results associated with different levels in one of the two predictors.

```{r}
library(rio)
elections2024 <- import("data/elections2024.csv")
demographics2024 <- import("data/demographics2024.xlsx")
elections2024<-elections2024 %>% left_join(demographics2024 %>% mutate(codecommune = as.numeric(codecommune), codedep = as.numeric(codedep)))

model3 <- lm(data = elections2024 %>% filter(dens == "Urbain intermédiaire") %>% 
               mutate(petr = etranger/pop, psup = sup/pop, pnodip = nodip/pop, log_revmoyfoy = log(revmoyfoy)),
             far_right_relvotes ~ log_revmoyfoy +petr+ pagri+pouvr+ pcadr+age+log(prixbien)+pempl+ppint+pindp+psup+pnodip + log_revmoyfoy*psup)

library(ggeffects)

ggpredict(model3, terms = c("log_revmoyfoy", "psup")) |> 
  as_tibble() |> 
  mutate(
    group = factor(
      group,
      levels = c("0.14", "0.23", "0.32"),
      labels = c(
        "Low share of higher education (14%)",
        "Medium share of higher education (23%)",
        "High share of higher education (32%)"
      )
    )
  ) |> 
  ggplot(aes(x, predicted, color = group, fill = group)) + 
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  labs(
    color = "Share with higher education",
    fill  = "Share with higher education"
  )
```
And you see that your predicted values change with the level of education : in the more educated cases, the line becomes less steep. Here, only three values of education are plotted by the program to give an idea of the interaction in three scenarios (commune with low/medium/high proportion of educated people), but this shows that the more educated people there are in a commune, the less the impact of income on the RN vote will be important in that commune. 

## Non-linear regressions

Linear regression relies on strong assumptions about the error term, in particular homoskedasticity and normality. However, there are many situations in which these assumptions are violated. A typical example arises when the outcome variable is binary, such as whether an individual voted in the last election ($Y = 1$ if yes, $Y = 0$ if no), and the explanatory variable is continuous, such as income.


If we attempt to fit a linear regression in this context, the resulting model exhibits strong heteroskedasticity, since the variance of the error term depends on the value of $X$. Indeed, when $Y$ is binary, the variance conditional on $X$ is given by $P(Y=1|X)(1 - P(Y=1|X))$, which varies with $X$. Moreover, the linear model may produce predicted values outside the logical range $[0,1]$, which is not meaningful when modeling probabilities.


```{r}
#| echo: false
# Load necessary library
# Load necessary library
library(ggplot2)

set.seed(123)

# Simulate continuous independent variable
n <- 300
x <- rnorm(n, mean = 0, sd = 1)

# Create probability using logistic function
true_prob <- 1 / (1 + exp(-( -1 + 2*x )))

# Generate binary outcome
y <- rbinom(n, size = 1, prob = true_prob)

data <- data.frame(x = x, y = y)

# Fit models
linear_model  <- lm(y ~ x, data = data)
logit_model   <- glm(y ~ x, data = data, family = binomial)

# Create prediction grid
x_seq <- seq(min(x), max(x), length.out = 200)
pred_data <- data.frame(x = x_seq)

pred_data$linear_pred <- predict(linear_model, newdata = pred_data)
pred_data$logit_pred  <- predict(logit_model, newdata = pred_data, type = "response")

# Plot
ggplot(data, aes(x = x, y = y)) +
  geom_jitter(height = 0.05, alpha = 0.4) +
  geom_line(data = pred_data, aes(y = linear_pred), 
            color = "red", size = 1.2) +
  labs(title = "Linear vs Logistic Regression for Binary Outcome",
       subtitle = "Red = Linear Regression | Blue = Logistic Regression",
       x = "Continuous Independent Variable",
       y = "Binary Outcome (0/1)") +
  theme_minimal()

```


Instead, a more appropriate approach is to directly model the conditional probability of the outcome given the explanatory variable. That is, we aim to estimate:

\begin{equation}
P(Y = 1 \mid X)
\end{equation}

One intuitive way to understand this is to divide the continuous variable $X$ into intervals and compute, within each interval, the proportion of observations for which $Y = 1$. This proportion provides an empirical estimate of the conditional probability of the outcome for values of $X$ in that interval.

In the figure, this idea is illustrated using several intervals of $X$. For each interval, the fraction of observations with $Y = 1$ is computed. We observe that this probability varies systematically with $X$, showing that the relationship between $X$ and $Y$ is not linear in levels, but instead affects the probability of the outcome.

This motivates the use of models specifically designed for binary outcomes, such as the logistic regression model, which ensures that predicted probabilities remain between 0 and 1 and properly accounts for the nonlinear relationship between $X$ and $P(Y=1|X)$.

The idea will instead be to associate every chunk of the continuous variables with probabilities of having, at this stage of the value $X$, hence to estimate:
\begin{equation}
P(Y = 1|X)
\end{equation}

Here, the example is shown for three different intervals of X : 

```{r}
#| echo: false
library(ggplot2)

set.seed(123)

# Simulate continuous independent variable
n <- 300
x <- rnorm(n, mean = 0, sd = 1)

# Create probability using logistic function
true_prob <- 1 / (1 + exp(-( -1 + 2*x )))

# Generate binary outcome
y <- rbinom(n, size = 1, prob = true_prob)

data <- data.frame(x = x, y = y)

# ---- Create 3 equal-width intervals of x ----
# ---- Create 15 equal-width intervals of x ----
breaks <- seq(min(data$x), max(data$x), length.out = 16)
data$x_group <- cut(data$x, breaks = breaks, include.lowest = TRUE)
# Fit models
linear_model  <- lm(y ~ x, data = data)
logit_model   <- glm(y ~ x, data = data, family = binomial)

# Create prediction grid
x_seq <- seq(min(x), max(x), length.out = 200)
pred_data <- data.frame(x = x_seq)

pred_data$linear_pred <- predict(linear_model, newdata = pred_data)
pred_data$logit_pred  <- predict(logit_model, newdata = pred_data, type = "response")

# Plot
ggplot(data, aes(x = x, y = y, color = x_group)) +
  geom_jitter(height = 0.05, alpha = 0.6) +
  geom_line(data = pred_data, aes(y = logit_pred), 
            color = "blue", size = 1.2) +
  labs(title = "Linear vs Logistic Regression for Binary Outcome",
       subtitle = "Red = Linear Regression | Blue = Logistic Regression",
       x = "Continuous Independent Variable",
       y = "Binary Outcome (0/1)",
       color = "X Interval") +
  theme_minimal()

```

```{r}
#| echo: false

library(ggplot2)
library(dplyr)

set.seed(123)

# Simulate continuous independent variable
n <- 3000
x <- rnorm(n, mean = 0, sd = 1)

# Create probability using logistic function
true_prob <- 1 / (1 + exp(-(-1 + 2*x)))

# Generate binary outcome
y <- rbinom(n, size = 1, prob = true_prob)

# Create data frame
data <- data.frame(x = x, y = y)

# ---- Create prediction grid ----
x_seq <- seq(min(x), max(x), length.out = 1000)
pred_data <- data.frame(x = x_seq)

# ---- Fit logistic regression ----
logit_model <- glm(y ~ x, data = data, family = binomial)

# ---- Predict logistic curve ----
pred_data$logit_pred <- predict(logit_model,
                                newdata = pred_data,
                                type = "response")

# ---- Create 15 equal-width intervals ----
breaks <- seq(min(data$x), max(data$x), length.out = 10)
data$x_group <- cut(data$x, breaks = breaks, include.lowest = TRUE)

# ---- Compute empirical proportions ----
group_summary <- data %>%
  group_by(x_group) %>%
  summarise(
    n_total = n(),
    n_ones  = sum(y == 1),
    n_zeros = sum(y == 0),
    prop_1  = n_ones / n_total,
    x_mid   = mean(x),
    .groups = "drop"
  )

# ---- Plot ----
# ---- Plot ----
ggplot(data, aes(x = x, y = y)) +
  
  geom_jitter(height = 0.05, alpha = 0.2, color = "gray") +
  
  # logistic regression curve
  geom_line(data = pred_data,
            aes(x = x, y = logit_pred),
            color = "black",
            size = 1.2) +
  
  # empirical proportions (colored by interval)
  geom_point(data = group_summary,
             aes(x = x_mid, y = prop_1, color = x_group),
             size = 3) +
  
  geom_line(data = group_summary,
            aes(x = x_mid, y = prop_1, color = x_group, group = 1),
            size = 1.2) +
  
  labs(title = "Logistic Regression vs Empirical Probabilities",
       subtitle = "Black = Logistic regression | Colored = empirical proportions per interval",
       x = "x",
       y = "Probability(y = 1)",
       color = "Interval") +
  
  theme_minimal()
```

The blue curve is called a logistic function. Mathematically, it specifies that the conditional probability follows a logistic function:

\begin{equation}
P(Y = 1 \mid X) = \frac{1}{1 + e^{-(\beta_0 + \beta_1 X)}}
\end{equation}

This functional form has several important properties. First, it ensures that predicted probabilities always lie between 0 and 1. Second, it allows for a nonlinear relationship between $X$ and the probability of the outcome. Third, it provides a convenient linear interpretation in terms of the log-odds:

\begin{equation}
\ln\left(\frac{P(Y = 1 \mid X)}{1 - P(Y = 1 \mid X)}\right) = \beta_0 + \beta_1 X
\end{equation}

Thus, logistic regression models the logarithm of the odds of the outcome as a linear function of the explanatory variable, while ensuring that predicted probabilities remain well-defined. The whole point of a logistic regression is therefore to estimate coefficients $\beta_0$ and $\beta_1$.

## Different logistic regressions
```{r}
library(ggplot2)

# Logistic function
logistic <- function(x, b0, b1) {
  1 / (1 + exp(-(b0 + b1 * x)))
}

# Create x grid
x <- seq(-5, 5, length.out = 400)

# Define parameter combinations
params <- data.frame(
  b0 = c(0, 0, -2, 2, 1),
  b1 = c(1, 2, 1, 1, -2),
  label = c("β0=0, β1=1 (baseline)",
            "β0=0, β1=2 (steeper)",
            "β0=-2, β1=1 (shift right)",
            "β0=2, β1=1 (shift left)",
            "β0=1, β1=-2 (shift left)")
)

# Generate predictions
plot_data <- do.call(rbind, lapply(1:nrow(params), function(i) {
  data.frame(
    x = x,
    prob = logistic(x, params$b0[i], params$b1[i]),
    label = params$label[i]
  )
}))

# Plot
ggplot(plot_data, aes(x = x, y = prob, color = label)) +
  geom_line(size = 1.2) +
  labs(title = "Effect of β0 and β1 in Logistic Regression",
       x = "X",
       y = "Predicted Probability",
       color = "Parameter Values") +
  theme_minimal() +
  ylim(0,1)

```

## Interpretation of logistic regression 

Interpreting the results of a logistic regression can initially feel challenging because the model predicts probabilities rather than direct outcomes. This may seem abstract at first. However, once you understand how to interpret the coefficients and odds ratios, it becomes much easier to see how the independent variables relate to the binary outcome—without necessarily transforming the coefficients into more intuitive values.

### The odds ratio

To properly understand the logit, we first need to clarify what it represents: it is the logarithm of the odds. The odds themselves are defined as the ratio between the probability that an event occurs, $p$, and the probability that it does not occur, $1-p$.

\begin{equation}
\text{ODDS} = \frac{p}{1-p}
\end{equation}

In other words, the odds tell us how much more (or less) likely the event is to happen than not to happen. For example, if an event has a probability of 0.1, its odds are:

\begin{equation}
\text{ODDS} = \frac{1/10}{9/10} = 1/9
\end{equation}

This means the event is nine times less likely to occur than not to occur.

If the probability is 0.8, then the odds are:

\begin{equation}
\text{ODDS} = \frac{8/10}{2/10} = 4,
\end{equation}

which means the event is four times more likely to happen than not to happen. Notice that when the probability equals 0.5—meaning the event is equally likely to occur or not—the odds equal 1.

Importantly, the odds are always positive and increase very rapidly as $p \rightarrow 1$. To obtain a more symmetric and analytically convenient measure, we usually take the natural logarithm of the odds. The logit is therefore defined as:

\begin{equation}
\ln(\text{ODDS}) = \ln\left(\frac{p}{1-p}\right)
\end{equation}

Unlike the odds, the logit can take any real value, positive or negative. The differences betweens the odds ratio function, the logit, and the logistic functions are shown in the panel below.

::: {.panel-tabset}

## Odds ratio
```{r}
p <- c(1:10000)/10000 # probabilities 0.0001, 0.0002, 0.0003,... 0.9998, 0.9999
plot(p, p/(1-p)) # odds
```
## Logit function
```{r}
plot(p, log(p/(1-p))) # logit
```
## Logistic function
```{r}
x <- c(1:10000)
mu <- 5000
s <- 500
plot(x, exp((x-mu)/s)/(1+exp((x-mu)/s))) # logistic function
```
:::


Now, suppose that the probability $p$ follows a logistic function of some variable $x$ (for instance, $p$ could represent the probability of voting, and $x$ a person’s income). Then we can write:

\begin{equation}
p(x) = \frac{1}{1 + e^{- (\beta_0 + \beta_1 x)}}
\end{equation}

Taking the logit of this expression yields a simple linear relationship:

\begin{equation}
\text{logit}(p(x)) = \beta_0 + \beta_1 x.
\end{equation}

This expression corresponds to the log of the odds. By exponentiating the logit, we recover the odds:

\begin{equation}
\text{ODDS} = e^{\text{logit}(p(x))} = e^{\beta_0 + \beta_1 x}.
\end{equation}

Therefore, a change $\delta x$ in $x$ leads to a multiplicative change of $e^{\beta_1 \delta x}$ in the odds. This makes interpretation straightforward: depending on the sign of $\beta_1$, the event becomes more or less likely to occur. 

In logistic regression, coefficients represent the change in the log-odds of the outcome for a one-unit increase in the predictor variable, while holding all other variables constant. The sign of the coefficient indicates the direction of the relationship:

- A positive coefficient means that as the predictor increases, the odds of the outcome increase.

- A negative coefficient means that as the predictor increases, the odds of the outcome decrease.

To make interpretation more intuitive, we often convert coefficients into odds ratios by exponentiating them (this simply reverses the logarithm). The odds ratio represents how the odds of the outcome change for a one-unit increase in the predictor variable compared to a reference value.

- An odds ratio greater than 1 indicates a positive association.

- An odds ratio less than 1 indicates a negative association.

It is also important to remember that logistic regression relies on certain assumptions, including linearity in the logit, independence of observations, and appropriate model specification. Violations of these assumptions can affect both model performance and interpretation. We will revisit diagnostic checks for logistic regression models in a later session.

## A real life example

## The ESS - European Social Survey

We will be working with the European Social Survey (ESS), the largest, most comprehensive, and arguably one of the most important surveys on social and political life within the European Union. The survey is conducted in biennial waves, and each participating European country that finances the survey generates its own data. For instance, the French data (which we'll be using from the latest, 10th wave) are produced at SciencesPo, specifically at the Centre de Données Socio-Politiques (CDSP)!

The ESS is an incredibly versatile resource, perfect for those who need a broad and detailed dataset for analyzing national politics in Europe or making cross-country comparisons. Learning how to handle, manage, and clean the ESS data will equip you with the skills to work with virtually any dataset you encounter. Additionally, many of you may find the ESS waves useful for your theses or research papers. The survey offers great potential for analysis, not just cross-sectionally but also over time. So, we encourage you to give it a try! 

## Importing the data

Of course, the first step will be for you to create an R project (.RProj), with a script for this class. Then we can load the library `tidyverse`. 

```{r}
library(tidyverse)
```

When importing a data set, you might need to make some changes in it, to prepare it for the analysis : these steps are essential before conducting any regression analysis (in fact, the regression stage is very rapid given the built-in functions in R). In fact, almost no data set you will download out there will be ready for your analysis. We will guide you through the basics of data management, to make sure you are able to transform data sets according to your needs. 

## Importing data

Here, I have already imported the data on my computer, and stored it in a subfolder called `data` located in the same folder as my script, which is why I call the relative path `data/ess_10.csv`. You can easily download the same data set by downloading it from the link we provide <>. You can either import it directly using `rio` and the `url` link we provide, or store it on your computer and import it using `import` or `read_csv`. A good reflex is to look at the variable names using `colnames()`, and have a glimpse at your data using `glimpse`, right after import, to inspect it.  

```{r}
library(rio)
ess <- import("https://github.com/scpo-quantimethods/quantitative_methods_two/raw/main/sessions/session2/data/ESS_10_fr.csv")
colnames(ess)
glimpse(ess)
```
This shows you that you have about 25 variables, some contain NA values (the data is clearly not cleaned yet). The variables refer to individual characteristics of the respondents, which you can find in the codebook of the ESS (provided when you download the data), either here <https://stessrelpubprodwe.blob.core.windows.net/data/round8/survey/ESS8_appendix_a7_e01_1.pdf>, or here <>.  Some variables are straightforward, with clear names like "essround" or "age," while others are less intuitive. Occasionally, we can infer the meaning of a variable from its name, but more often than not—either because the name is confusing or the abbreviation doesn't make much sense—we need to refer to the dataset’s documentation. For this particular version of the dataset, you'll find the documentation in an HTML file on Moodle (session 2).

Every high-quality, reliable dataset should have some form of documentation. If it doesn't, it's a red flag, and I would advise caution before using it! This documentation is typically called a "codebook." Codebooks can vary greatly in quality—some are well-organized and easy to read, while others are a challenge to navigate. In this class, you’ll encounter both types, so you'll be prepared to work with whatever you come across.

In general, data set you will download online will come with a codebook, basically associating each variable name with its actual name. In the case of variable `gndr`, the codebook for instance informs you that it means gender, and that the values refer to: 

|Value| 	Category|
|-----|-----------|
|1 	|Male|
|2 	|Female|
|.a (9)| 	No answer|

Moreover, you see that you have 33 351 observations - corresponding to the number of respondents of the ESS. 

In fact, this dataframe still includes many variables that we won’t need later or that contain no useful information. Let's remove these first. While this step can be done later, it's often a good idea to tidy up the dataset at the very beginning, so you start with a cleaner, more manageable dataset.

You can select the variables you need (using `select()`) right when importing the CSV file. You can remove variables by calling `select(-c(your_variable_name1,your_variable_name2))`.
```{r}
ess <- ess  |>
  dplyr::select(
    cntry,
    polintr,
    trstplt,
    trstprt,
    vote,
    prtvtefr,
    clsprty,
    gndr,
    yrbrn,
    eduyrs,
    emplrel,
    uemp12m,
    uemp5yr
  )
colnames(ess)
unique(ess$cntry)
table(ess$cntry)
```
This shows you . The functions `unique` and `table` allow you to analyse the distribution of a categorical variables, among its different values. You see that France has 1977 observations, which means that the sample contains about 1977 French respondents. You can restrict the data set to the French values by simply calling `filter()`, which will basically only keep the values that satisfy a certain condition (here `cntry == "FR"`), and also use table to analyse categorical values with numeric values, like vote.

```{r}
ess_fr <- ess %>% filter(cntry == "FR")
table(ess_fr$vote)
```

If you look at the codebook online, you will find that the values correspond to the following categories : 

|Value| 	Category|
|-----|-----------|
|1 	|Yes|
|2 	|No|
|3 	|Not eligible to vote|
|.a (7) 	| 	Refusal*|
|.b (8) 	| 	Don't know*|
|.c (9)	|No answer*|

And that the question that was asked to the respondents was "some people don't vote nowadays for one reason or another. Did you vote in the last [country] national election in [month/year] ?".

Hence, you see that 590 respondents did not vote in the last election, 1025 did and 307 were not eligible to vote.

The values 7, 8, and 9 are quite common in survey datasets. They typically represent situations where respondents either didn't want to answer, their answer was unclear, or they couldn't remember. You'll encounter these values in almost every survey-based dataset.

As already said, we'll focus on voter turnout. Our goal is to understand what factors encourage people to vote and what decreases the likelihood of voting on election day. Therefore, the question for our dependent variable will be: Did the respondent vote or not? Mathematically, this question cannot be answered using linear regression with the OLS method, as the dependent variable is binary—0 represents "did not vote" and 1 represents "voted." There are only two possible outcomes, and the variable is not continuous, which violates one of the assumptions of OLS.

However, if we use the voting turnout variable as it is right now, we won't have a valid binary variable. It also won't be reliable, as it includes values that are either irrelevant or will skew our estimations. In fact, logistic regression (logit) can only be applied to binary variables.

So, before we proceed, we need to transform our dependent variable. We'll remove unwanted values and convert the 1s and 2s into 0s and 1s.

First, we need to eliminate the values we’re not interested in. While it's possible to filter out these values directly (as shown below), I recommend against it. Instead, you should transform these values into NA. Rows with NA values will be excluded from the regression, but if you later decide to run another model with a different dependent variable, you may lose valuable observations that you've filtered out. Therefore, a good rule of thumb is to only filter out values if you're certain you’ll never need them again.

```{r}
ess <- ess |>
  filter(!vote %in% c(3, 7, 8, 9))
```

We can directly mutate the values within the variable using `case_when()` inside the `mutate()` function. While `case_when()` may seem a bit tricky to read at first, it’s a powerful tool. Essentially, `case_when()` evaluates a series of conditions and applies the corresponding transformation to each element of a vector or column in a dataset.

Within the function, you must specify at least one condition, and you can include multiple conditions if needed. If you want the conditions to be evaluated simultaneously, use the & operator between them. If you want the condition to apply when either of two conditions is true, you would use the `|`` operator. After specifying the conditions, you use a tilde `~`, followed by the new value that the variable should take on if the condition is met.

Here’s how it works:

```{r}
#| eval: false
case_when(
  condition1 ~ value_if_true1,
  condition2 ~ value_if_true2,
  TRUE ~ default_value
)
```

Let’s look at an example below. I want the values 3, 7, 8, and 9 in the vote variable to be replaced with NA, as they are irrelevant to our analysis. To do this, we use the mutate() function, and I assign a new column (which I’ll name vote, the same as the original variable). Inside the case_when() function, I specify vote %in% c(3, 7, 8, 9) ~ NA_integer_. This statement means: whenever vote contains the values 3, 7, 8, or 9, replace them with NA_integer_.

The next line, TRUE ~ vote, serves as a catch-all rule. It says, "For all other cases, keep the original values of vote." This ensures that any value not matching the previous condition remains unchanged.

Here’s the code:
```{r}
ess <- ess |> 
  mutate(
    vote = case_when(
      vote %in% c(3, 7, 8, 9) ~ NA_integer_,
      TRUE ~ vote
  ))
table(ess$vote)
```

Now, we just need to transform the ones and twos into zeros and ones. This is both a matter of convention, my personal coding preference, and, of course, to give you a bit more practice with data management. Since we are focusing on people who have voted, we’ll code those who voted as 1 and those who didn’t as 0.

We’ll use `case_when()` again to do this. The code `case_when(vote == 1 ~ 1, vote == 2 ~ 0)` works as follows: whenever vote equals 1 (indicating that the person voted), we keep it as 1; whenever vote equals 2 (indicating that the person did not vote), we assign it 0.


With that, we now have our binary dependent variable, which is essential for performing a logit. The `table()` function will show that we’ve successfully changed the values of the variable, but the total count remains the same.

In Base R, you could achieve this in a different way, but I personally find that using the `mutate()` function is the most elegant approach. That said, it's entirely up to you which method you prefer.

```{r}
# only use the ifelse() function
ess$vote <- ifelse(ess$vote == 1, 1, 0)

# This will leave the value of 1 at 1 and change 2 to 0 for the column vote.
ess$vote[ess$vote == 1] <- 1
ess$vote[ess$vote == 2] <- 0
```

We’re very close to finishing the data management part and getting ready to work on our model. However, there are still many variables that are as "untidy" as our initial dependent variable. We need to check the codebook to see if there are specific values we can’t simply replace across the entire dataset and whether they are still irrelevant to our analysis.

For our independent variables of interest, here's what we need to do:

- **Political interest (`polintr`)**: Remove values c(7:9).
- **Trust in politicians (`trstplt`)**: No recoding needed (unwanted values are already `NA`).
- **Trust in political parties (`trstprt`)**: Already done.
- **Feeling close to a party (`clsprty`)**: Convert 1/2 to 0/1 and remove values c(7:8).
- **Gender (`gndr`)**: Convert to 0/1 and remove "no answer" values.
- **Year born (`yrbrn`)**: Already done.
- **Years of full-time education completed (`eduyrs`)**: Already done.

We’ll tackle all these steps at once using pipes `%>%` (from the `tidyverse`) or `|>` (Base R), which allows us to perform everything in a neat and efficient way. I’ll also re-run the mutation for our dependent variable to demonstrate the flexibility of long pipe chains.

We start by using `mutate()` to adjust the independent variables we need for our model. Notice how you can separate the different manipulations with commas inside the same function, which results in clean, readable code. There are various functions for data wrangling, such as `recode()`, `if_else()`, and `case_when()`. For simplicity and clarity, I’ll be using `case_when()` here, though there’s a newer function, `case_match()`, that you could also use. 

Now, let's talk about `NA_integer_`. In R, `NA` signifies "not available" or missing data. But R is very particular about the type of data it's working with. If your variable is numeric, you need to specify that missing values should be treated as numbers. By using `NA_integer_`, you're telling R, "This is missing data, but if it were here, it would be a number." This ensures consistency and avoids confusion when R processes your data. If you're working with character strings, you'd use `NA_character_`. For numeric values with decimals, you’d use `NA_real_` instead of `NA_integer_`.

```{r}
ess_final <- ess |> 
  # mutating the dependent variable to get rid of any unnecessary rows
  # as we have done above
  mutate(
    vote = case_when(
      vote %in% c(3, 7, 8, 9) ~ NA_integer_,
      vote == 2 ~ 0,
      vote == 1 ~ 1, 
      TRUE ~ vote
    ),
    # Recode 'polintr' to drop values 7 to 9
    polintr = case_when(
      polintr %in% c(7:9) ~ NA_integer_,
      TRUE ~ polintr
    ),
    # Recode 'clsprty' 1/2 into 0/1 and drop 7, 8
    clsprty = case_when(
      clsprty == 1 ~ 0,
      clsprty == 2 ~ 1,
      clsprty %in% c(7, 8) ~ NA_integer_,
      TRUE ~ clsprty
    ),
    # Recode 'gndr' into 0/1 and handle other cases as NAs
    gndr = case_when(
      gndr == 1 ~ 0,
      gndr == 2 ~ 1,
      TRUE ~ NA_integer_
    ),
    # recode eduyrs
    eduyrs = case_when(
      eduyrs %in% c(1:55) ~ eduyrs,
      TRUE ~ NA_integer_
    ),
    # this one is very advanced...
    across(c(trstplt, trstprt), ~case_when(
      .x <= 10 ~ .x, 
      TRUE ~ NA_integer_
    ))
  )
```

## Running the model 

Once the data preparation is done, the best part is that running a logistic regression model literally takes just one line (or two, depending on the number of independent variables). As you’ll see, it’s quite straightforward.

The `glm()` function, which stands for **generalized linear model**, is a part of Base R. 

In Professor Rovny's lecture, we discussed that for Maximum Likelihood Estimation (MLE), you need to know or assume the distribution of your dependent variable. Based on that distribution, you choose the appropriate linear model. For a binary outcome like voting (yes/no), the distribution is **binomial**. So, within the `glm()` function, we specify the distribution family as binomial. 

It's worth mentioning that you can specify other families, such as **Gaussian**, **Poisson**, or **Gamma**, depending on your data. We won’t go into detail about those families here, but the `glm()` function is quite powerful. By setting `family = binomial(link = "logit")`, we specify that we're conducting a **logistic regression**. If you wanted to use a different link, such as **probit** or **cauchy**, you can easily adjust it here. The default is set to logit, so in most cases, you won’t need to specify it explicitly. 

If your dependent variable is incompatible with the chosen family, the function will alert you to a problem.

For the model we’re building, we're assuming that voting behavior (whether a person voted or not) is influenced by political interest, trust in politicians, trust in political parties, and feeling close to a party, along with typical control variables like gender, age, and education. 

Please note that this regression isn’t sophisticated enough to be published—it’s simply an example. My hypothesis is that factors like political interest, trust in politics and politicians, and party affiliation are key explanatory variables.

Just like with an OLS regression, we start by specifying the dependent variable (in this case, **vote**). Then, we separate the dependent variable from the independent variables with a `~`. Next, we list the predictors, separated by a `+`. Finally, we specify the data object and the model family.

Here’s how you can run the model:

```{r}
logit <- glm(
  vote ~ polintr + trstplt + trstprt + clsprty + gndr +
    yrbrn + eduyrs,
  data = ess_final,
  family = "binomial")
```

The object we created, called **logit**, now contains our model along with its coefficients, confidence intervals, and many other details that we’ll explore further! As you can see, the actual process of constructing the model is incredibly simple and only requires a single line of code. 

Now that we've created the model, we can examine its components and assess how well it fits our data. The logistic regression model has produced outputs that we can use to make inferences about the relationship between our dependent and independent variables. The simplicity of the code doesn’t mean that the analysis is complete—it’s just the beginning of our exploration!

```{r}
library(broom)
tidy(logit)
```


```{r}
stargazer::stargazer(
  logit,
  type = "text",
  dep.var.labels = "Voting Behavior",
  dep.var.caption = c("Voting turnout; 0 = abstention | 1 = voted"),
  covariate.labels = c(
    "Political Interest",
    "Trust in Politicians",
    "Trust in Parties",
    "Feeling Close to a Party",
    "Gender",
    "Year of Birth",
    "Education"
  ),
  star.cutoffs = c(0.05, 0.01, 0.001)
)
```


We saw that a change $\delta x$ in $x$ leads to a multiplicative change of $e^{\beta_1 \delta x}$ in the odds. This makes interpretation straightforward: depending on the sign of $\beta_1$, the event becomes more or less likely to occur. 

Hence, in our case : 

```{r}
# simply use exp() on the coefficients of the logit
exp(coef(logit))
# here would be a second way of doing it
exp(logit$coefficients)
```

We can—and should—also report the 95% confidence intervals (CIs). As a brief reminder, a confidence interval provides a range of values that is likely to contain the true value of a parameter—in our case, the coefficients of the predictor variables. This range is associated with a specific level of confidence. The most commonly used confidence levels (note that this is simply a statistical convention) are 95% and, less frequently, 99%.

A 95% confidence interval means that if we were to repeatedly estimate the same logistic regression model using many different samples drawn from the same population, the true parameter value would lie within the calculated confidence interval in 95% of those samples.

```{r}
# most of the times the extra step in the next lines is not necessary and this 
# line of code is enough
exp(cbind(
  OR = coef(logit), 
  confint(logit)))
```

Once we exponentiate the coefficient and obtain the odds ratio (OR), interpretation becomes more straightforward. The odds ratio tells us how the odds of the outcome change with a one-unit increase in a predictor variable. For example, if the odds ratio for political interest is below 1, this means that a one-unit increase in political interest decreases the odds of voting (compared to not voting). The same logic applies to the other variables in the model.

One final point about odds ratios—arguably the most intuitive way to interpret them—is to translate them into percentages.

```{r}
100*(exp(logit$coefficients[-1])-1)
```

## The logistic regression 

In logistic regression, coefficients represent the change in the log-odds of the outcome for a one-unit increase in the predictor variable, while holding all other variables constant. The sign of the coefficient indicates the direction of the relationship:

- A positive coefficient means that as the predictor increases, the odds of the outcome increase.

- A negative coefficient means that as the predictor increases, the odds of the outcome decrease.

To make interpretation more intuitive, we often convert coefficients into odds ratios by exponentiating them (this simply reverses the logarithm). The odds ratio represents how the odds of the outcome change for a one-unit increase in the predictor variable compared to a reference value.

- An odds ratio greater than 1 indicates a positive association.

- An odds ratio less than 1 indicates a negative association.

It is also important to remember that logistic regression relies on certain assumptions, including linearity in the logit, independence of observations, and appropriate model specification. Violations of these assumptions can affect both model performance and interpretation. We will revisit diagnostic checks for logistic regression models in a later session.

I know this may sound dense. It is somewhat technical, but it becomes much clearer once we apply the concepts to our actual model. Interpreting exponentiated coefficients as odds ratios will also make the results much more intuitive.

Looking at our model summary, we can already see that most independent variables are statistically significant (p < .05). Typically, we focus our interpretation on statistically significant variables and do not interpret those that are not significant.

From the results, we observe that closeness to a party and age are negatively associated with voting behavior. This makes sense given how the variables are coded. According to the codebook, higher values represent respondents who are less interested, less close to a party, or older (depending on the coding). Therefore, higher values correspond to a lower likelihood of voting on election day.

Trust in politicians, however, is coded in the opposite direction. Higher values indicate lower trust. As a result, the positive coefficient suggests that lower trust in politicians increases the likelihood of not voting. Ideally, it would have been better to code all independent variables in the same direction for easier interpretation, but the results can still be interpreted correctly as long as we carefully consider the coding scheme.

## Predicted probabilities 

Overall, we could now present a table with the odds ratios and move on to writing the paper. However, visualizing results often tells a much clearer and more compelling story. In this section, I will show you how to compute predicted probabilities and how to display them in an effective and visually appealing way.

In logistic regression, predicted probabilities represent the likelihood that an event occurs given specific values of the predictor variables. Rather than interpreting log-odds or odds ratios—which can feel abstract—predicted probabilities provide a simple number between 0 and 1. For instance, instead of stating that higher political interest increases the log-odds of voting, we can say that an individual with high political interest has a 75% probability of voting. This makes the results far more intuitive and easier to communicate.

Predicted probabilities are expressed on the original scale of the independent variables (IVs). This means they directly illustrate how changes in variables—such as political interest, trust in politicians, or education—affect the probability of the outcome. If political interest is measured on a scale from 1 to 4, we can examine how the probability of voting changes at each level of that scale, rather than simply noting whether the relationship is positive or negative. This is a major advantage over coefficient plots, which show direction and magnitude but not the actual likelihood of the outcome. Predicted probabilities therefore allow us to present findings in a way that is both intuitive and closely tied to the real-world measurement of our variables.

The key idea is that we select one predictor variable and decide the values at which we want to estimate the outcome. While doing so, we “hold the other independent variables constant.” This allows us to isolate the effect of the variable we are interested in.

To make this absolutely clear: when calculating predicted probabilities, we focus on one independent variable of interest and keep all other variables fixed—typically at their mean or median, or, in the case of a dummy variable, at one of its two possible values. By holding the other variables constant, we can clearly observe the independent effect of the predictor we care about.


```{r}
# this package contains everything we need craft predicted probabilities and
# visualize them as well
library(ggeffects)

# like the predcit() function of Base R, we use ggpredict() and specify
# our variable of interest
predicted_probs <- ggpredict(logit, terms = "polintr")
predicted_probs

# this is the simplest way of plotting this
ggplot(predicted_probs, aes(x = x, y = predicted)) +
  # our graph is more or less a line, so geom_line() applies
  geom_line(color = "blue", size = 1) +
  geom_point() +
  # geom_ribbon() with the values that ggpredict() provided for the confidence
  # intervals then gives
  # us a shade around the geom_()line as CIs
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), fill = "grey", alpha = 0.3) +
  labs(
    title = "Predicted Probability of Voting by Political Interest",
    x = "Political Interest",
    y = "Predicted Probability of Voting"
  ) + 
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )

```

```{r}
predicted_probs_gender <- ggpredict(logit, terms = "gndr") |> 
  mutate(
    gender = x |> as.factor()
  ) |> 
  na.omit()

ggplot(predicted_probs_gender,
       aes(x = gender, y = predicted, fill = gender)) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  geom_point(shape = 21, size = 5) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Predicted Probability of Voting by Gender",
       x = "Gender",
       y = "Predicted Probability of Voting",
       fill = "Gender") +
  scale_x_discrete(drop = TRUE) +
  theme_minimal()
```


## The maximum of likelihood estimation 

### Constant probabilities 

Suppose you are interested in the outcome of a binary event, where the probability of the event occurring is $p$ and the probability of it not occurring is $1 - p$. This follows Bernoulli’s probability distribution:

\begin{equation}
P(X = x) = p^x (1-p)^x \ \text{with} X \in \{0, 1\}
\end{equation}

For example, imagine rolling a fair six-sided die. The probability of rolling a 1 (i.e., $X = 1$) is $p = 1/6$, and the probability of rolling anything other than a 1 (i.e., $X = 0$) is $1 - p = 5/6$. Many research questions involve finding such probabilities for binary outcomes, such as understanding why people vote or abstain, or why they vote for a particular party. These are binary outcomes, and calculating their probabilities using linear regression would not be appropriate.

Suppose you observe $N$ independent binary outcomes, each determined by an unknown probability $p$. Let $n_1$ represent the number of outcomes where $X = 1$, and $N - n_1$ the number of outcomes where $X = 0$. The likelihood function is:

\begin{equation}
\mathcal{L}(p) = \prod_i^{n_1} (p_i) \prod_i^{N-n_1} (1-p_i) = p_i^{n_1}(1-p_i)^{N-n_1}
\end{equation}

This expression represents the probability of obtaining $X = 1$ $n_1$ times, multiplied by the probability of obtaining $X = 0$ $N - n_1$ times. If you do not know the probability $p$ but have the outcomes for a large number of observations, you can estimate $p$ by maximizing the likelihood function. This method is called \textbf{maximum likelihood estimation}.

### A simple example
Let’s say you're interested in estimating the probability of rolling a 1 on a fair die, but you don't know the probability (which you know theoretically to be 1/6). Let $X = 1$ if you roll a 1 and $X = 0$ otherwise. The probabilities are:

\begin{equation}
P(X=1) = 1/6 = p_1
\end{equation}

and
\begin{equation}
P(X=0) =5/6 = p_0 = 1-p_1
\end{equation}

You roll the die 10,000 times and want to estimate $p_1$. The likelihood function for these observations would be:

\begin{equation}
\mathcal{L} = \Pi_i^{n_1} (p_i) \Pi_i^{10000-n_1} (1-p_i)
\end{equation}

You would typically compute the likelihood function as:
\begin{equation}
\mathcal{L}(p) = \Pi_i^{n_1} (p_i) \Pi_i^{10000-n_1} (1-p_i) = p_i^{n_1} (1-p_i)^{10000-n_1}.
\end{equation}
The goal is to maximize the likelihood function to find the value of $p$ that makes the observed data most probable. To do this, take the log of the likelihood function:

\begin{equation}
\text{log}(\mathcal{L}(p)) = \text{log}(\Pi_i^{10000} (p_i) \Pi_i^{10000} (1-p_i)) = \text{log}(p_i^{N_1})+\text{log}((1-p_i)^{10000-N_1
\end{equation}

which gives : 

\begin{equation}
 \text{log}(p_i^{N_1})+\text{log}((1-p_i)^{10000-N_1}) = N_1 \text{log}(p_i) + (10000-N_1)\text{log}(1-p_i)
\end{equation}

Next, differentiate with respect to $p$:
\begin{equation}
\frac{d\text{log}(\mathbb{L}(p))}{dp} = \frac{n_1}{p} - \frac{10000 - n_1}{1 - p}.
\end{equation}
Set the derivative equal to zero to solve for $p$:
\begin{equation}
\frac{n_1}{p} = \frac{10000 - n_1}{1 - p}.
\end{equation}

This simplifies to:
\begin{equation}
p = \frac{n_1}{10000}.
\end{equation}

This result is simply the proportion of 1's in the 10,000 rolls, which is the maximum likelihood estimate of the probability. According to the law of large numbers, as $N \to \infty$, this estimate will converge to the true probability $p$. Thus, maximizing the likelihood function allows us to estimate the probability distribution that best fits the data.

## Non-constant probabilities 

Now imagine that your probabilities depend on other explanatory variables $x_1, x_2, x_3$. For example, in a research study seeking to understand voter behavior, the probability of an individual voting might depend on their income, gender, or political affiliation. In this case, you would seek to estimate the relationship between these variables and the outcome of voting.

The likelihood function in this case would be:

\begin{equation}
L(x_1, x_2, x_3) = \Pi_{i\  (y_i = 1)}^n p(x_1, x_2, x_3)_i \Pi_{i\  (y_i = 0)}^n (1 - p(x_1, x_2, x_3))_i
\end{equation}

In the simpler case where only one explanatory variable $x$ is considered, the likelihood function is:

\begin{equation}
L(x) = \Pi_{i\  (y_i = 1)}^n p(x)_i \Pi_{i\  (y_i = 0)}^n (1 - p(x)_i)
\end{equation}

which you would maximize to find the parameters that best fit the data.

## Logistic function

The logistic function is :
\begin{equation}
p_L (x) = \frac{1}{1+ e^{-(\beta_0 + \beta_1 x)}} = \frac{1}{1+e^{-(x-\mu)/s}} = \frac{e^{(x-\mu)/s}}{1+e^{(x-\mu)/s}}
\end{equation}
and it is strictly bounded between 0 and 1, which is expected for a probability.

```{r}
x <- c(1:10000)
mu <- 5000
s <- 500
plot(x, exp((x-mu)/s)/(1+exp((x-mu)/s)))
```

You can use this logistic function in the likelihood function:
\begin{equation}
\mathcal{L}(x) = \Pi_{i\  (y_i = 1)}^n p(x)_i \Pi_{i\  (y_i = 0)}^n (1 - p(x)_i),
\end{equation}
maximize it, and find the parameters $\beta_0$ and $\beta_1$ that best fit your data.

This is the logic behind maximum likelihood estimation: you assume your probability follows a certain distribution (in this case, the logistic distribution), and then estimate the parameters that maximize the likelihood of obtaining your data. This is precisely what the `glm` function in R does when it fits a logistic regression model — it computes the parameters that maximize the likelihood function using numerical methods.

## Logit
The logit is the inverse of the logistic function. It is defined as:
\begin{equation}
\text{logit}(p(x)) = \ln (\frac{p(x)}{1-p(x)}) = \beta_0 + \beta_1 x,
\end{equation}

The logit is the inverse of the logistic function. It is defined as:

\begin{equation}
\text{logit}(p(x)) = \ln \left( \frac{p(x)}{1-p(x)} \right) = \beta_0 + \beta_1 x,
\end{equation}

In other words, the logit is the logarithm of the odds. The odds is simply the ratio ($ \frac{p}{1-p}$ ), and gives a real number that tells you how many times more likely (or less likely) you are to obtain the outcome. For example, if the probability of an event occurring is 20% (i.e., ($ p = 0.2 $)), the odds are:
\begin{equation}
\text{odds} = \frac{0.2}{1 - 0.2} = 0.2 / 0.8 = 0.25.
\end{equation}
This means the event is 4 times less likely to happen than not to happen.

On the other hand, if the probability of an event occurring is 80% (i.e., ($ p = 0.8 $)), the odds are:
\begin{equation}
\text{odds} = \frac{0.8}{1 - 0.8} = 0.8 / 0.2 = 4.
\end{equation}
This means the event is 4 times more likely to happen than not to happen.

When considering probabilities between 0 and 1, the odds function ($\frac{x}{1-x}$) increases very rapidly as the denominator approaches zero. This is why the log of the odds (the logit) provides a more manageable and continuous scale. For a visual comparison, here's a plot in R showing the relationship between the probability ($p$) and the odds, as well as the logit : 

```{r}
p <- c(1:10000)/10000
plot(p, p/(1-p))
plot(p, log(p/(1-p)))
```
Hence, you see that the logit associates every probability $p$ with a continuous variable. In the special case where the probability $p$ is a logistic function of variables $X_1, X_2, \dots, X_k$, we can express this relationship as:

\begin{equation}
\text{logit}(p | x_1, x_2, x_3) = \ln \left( \frac{p}{1 - p} \right) = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_3.
\end{equation}

This is why you can obtain the odds by simply exponentiating the logit, which essentially gives you the odds ratio. In other words:

\begin{equation}
\text{ODDs} = \frac{p}{1 - p} = e^{\text{logit}(p)} = e^{\beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_3}.
\end{equation}

Any variation $\delta x_1$ in the explanatory variable $x_1$ leads to a change in the odds:

\begin{equation}
\delta \text{ODDs} = e^{\beta_1 \delta x_1}.
\end{equation}

This means that any change in $x_1$ results in a multiplicative change in the odds of the outcome occurring. The model allows you to compute the probability that best characterizes your process and helps to understand how each explanatory variable influences the outcome.

Now, if you decide that $p$ is a function of other variables $x_1, x_2, x_3$, you need a function that associates these real numbers with probabilities that range from 0 to 1. This is exactly what the logistic regression model does — it provides a way to model the relationship between the independent variables and a binary outcome, ensuring that the probabilities stay within the valid range of 0 and 1.

But now, imagine you want to understand what drives the probability. In other words, you want to know how different factors (i.e., the explanatory variables) affect the likelihood of the outcome. The logistic regression model, through maximum likelihood estimation, gives you the answer to this question by estimating the coefficients $\beta_1, \beta_2, \dots, \beta_k$, which tell you the effect of each variable on the log-odds of the outcome.

