---
title: "Methods for Research and Policy Evaluation"
subtitle: "Session 3"
author: "Paul Servais"
format: revealjs
---


## We saw in previous sessions
- How to describe variables : descriptive statistics
- How to infer from sample to larger population : statistical inference
- How to analyse relations between variables : 
  - When DV is continuous:
    - Indep. variable = categorical : t-test
    e.g. Urban/rural on median income
    - Indep. variable = continuous : linear regressions
    e.g. Far-right vote share in communes and income
    
## Summary 
- You can have different types of independent and dependent variables

|                         | **DV: Categorical**              | **DV: Interval/Continuous**          |
|-------------------------|-----------------------------------|----------------------------------------|
| **IV: Categorical**                | **Crosstabs / χ² Test**          | **Compare Means / t-Test**             |
| **IV: Interval/Continuous**        | **Logistic Regression / Logits** | **Correlation / Scatter / Regression**  |

- Today : Correlations, and logistic regressions

# Yet we did not discuss
- Under which assumptions is OLS reliable ?
- Can independent variables interact between each other ?
- What if the dependent variable is not continuous ?
  - e.g. many policy outcomes are binary : 
    - did the policy increase school enrollment ?
    - did the policy increase program participation ?
    - did the policy decrease hospital admissions ?

## Content

- Theory : 
  - OLS regressions : diagnostics 
  - Interactions in linear regressions
  - Non-linear regressions (logit/logistic regressions)
- Exercices

## Objectives 
- For you to know the limitations of linear regressions
  - multicollinearity
  - heteroskedasticity 
  - endogeneity
- For you to know about alternative regression models
  - logits, probits
  - be able to interpret the regression coefficients 

# OLS regressions : diagnostics 

## The linear regression {.scrollable}
- We saw :
\begin{equation}
Y_i = \beta_0 + \beta_1 X_i + \epsilon_i
\end{equation}
- Note about notations
- Cross section :
\begin{equation}
Y_i = \beta_0 + \beta_1 X_i + \epsilon_i
\end{equation}
- Time series data :
\begin{equation}
Y_t = \beta_0 + \beta_1 X_t + \epsilon_t
\end{equation}
- Panel data :
\begin{equation}
Y_{it} = \beta_0 + \beta_1 X_{it} + \epsilon_{it}
\end{equation}



- Coming back to our example
```{r}
#| echo: true
library(ggplot2)
library(tidyverse)
library(rio)

elections2024 <- import("data/elections2024.csv")
socio_demo <- import("data/demographics2024.xlsx")
elections2024 <- elections2024 %>% left_join(socio_demo %>% mutate(codecommune = as.numeric(codecommune), codedep = as.numeric(codedep)))
```

## Simple linear regression model {.scrollable}

```{r}
#| echo: true
#| eval: true
library(ggplot2)
elections2024 %>% 
  filter(dens == "Urbain intermédiaire") %>% 
  ggplot(aes(x = log(revmoyfoy), y = far_right_relvotes)) + geom_point() + labs(y = "RN vote", x = "Median income in commune (log scale)") + geom_smooth(method = "lm") +theme_minimal()
```

## Minimising the errors 

```{r}
library(ggplot2)
library(dplyr)
library(ggplot2)

df_plot <- elections2024 %>% 
  filter(dens == "Urbain intermédiaire") %>% 
  mutate(x = log(revmoyfoy)) %>% 
  mutate(
    y_hat = predict(
      lm(far_right_relvotes ~ x, data = .)
    )
  )

ggplot(df_plot, aes(x = x, y = far_right_relvotes)) +
  geom_point() +
  geom_segment(
    aes(
      xend = x,
      yend = y_hat
    ),
    color = "red",
    alpha = 0.5
  ) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    y = "RN vote",
    x = "Median income in commune (log scale)"
  ) +
  theme_minimal()
```

## Interpretation of linear regression in R
- Intercept : predicted value when predictors = 0
- Estimate  
  - sign
  - magnitude 
- Significance levels : compares with $\beta_1 = 0$
  - t-value ($df>1000$) : $t_{.90} = 1.645$, $t_{.95} = 1.96$, $t_{.99} =2.58$
  - p-value : probability that your result is due to NULL


## The $R^2$ value
- You can divide your explained variance by your total variance:
\begin{equation}
R^2 = \frac{\text{ESS}}{\text{TSS}} = \frac{\sum_i^N (\hat{Y}_i - \bar{Y})^2}{\sum_i^N (Y_i - \bar{Y})^2}
\end{equation}
- A higher $R^2$ indicates how much of the variance is explained by the model


## There are five assumptions 

* Exogeneity : $\mathbb{E} (\epsilon_i|X_i) = 0$. 
* Homoskedasticity : $\mathbb{E} (\epsilon_i^2) = \sigma^2$.
* $\mathbb{E} (\epsilon_i \epsilon_j)= 0$ or $\mathbb{C}ov(\epsilon_i, \epsilon_j) = 0$
* $X_{ij}$ is non random $\forall  i = 1,...,n ; j = 1, ..., k$ (X is not a random variable, contrary to the error and the dependent variable).
* Multicollinearity : $\text{rank}((X_{ij})) = k$. In the econometric jargon, violating that assumption would lead to a model with so-called *multicollinearity* between predictors.

## Take a simple example {.scrollable}

- You want to verify the physical linear law :
\begin{equation}
W = m g
\end{equation}
- You ask 100 people to measure the weight of 100 copper masses of 0.1 to 10kg.
- Your model would be :
\begin{equation}
W_i = g m_i + \epsilon_i
\end{equation}

```{r}
#| echo: true
#| eval: false
library(tidyverse)
library(tidyverse)

m <- sample(1:100, 100, replace = FALSE) / 10
W <- 9.81 * m + rnorm(100, 0, 15)

dat <- data.frame(m, W)
lm_model_biased <- lm(W ~ m, data = dat)
```

## You might get something like...

```{r}
#| echo: false
#| eval: true
library(tidyverse)
library(tidyverse)

m <- sample(1:100, 100, replace = FALSE) / 10
W <- 9.81 * m + rnorm(100, 0, 15)

dat <- data.frame(m, W)
lm_model_biased <- lm(W ~ m, data = dat)

dat %>% 
  ggplot(aes(x = m, y = W)) + 
  geom_point() +
  geom_smooth(method = "lm",
              linetype = "dashed",
              color = "black",
              se = TRUE,
              linewidth = 1,
              alpha = 0.4)+
  theme_minimal()
```

## And your estimates
```{r}
summary(lm_model_biased)
```

## Yet other factors can influence the precision of the estimate
- The calibration of the measuring instrument
- The individual abilities of the people measuring
- etc. 
- Exogeneity = once you control for mass, the expectation of the error is zero
- In other words, the error is not related to mass in a systematic way

## Endogeneity
- Imagine the case where your weighing machine is sensitive to mass
- the precision decreases with $m^3$ : much less precise for larger masses
\begin{equation}
E(\epsilon|X) =/= 0
\end{equation}
- Important bias
- Also appearing with omitted variable bias (e.g. income and education on far-right vote)

## Endogeneity {.scrollable}
```{r}
#| echo: false
#| eval: true
library(tidyverse)
library(tidyverse)

m <- sample(1:100, 100, replace = FALSE) / 10
W <- 9.81 * m + rnorm(100, m**3, 15)

dat <- data.frame(m, W)
lm_model_biased <- lm(W ~ m, data = dat)
summary(lm_model_biased)

```
- Important bias

## Endogeneity 

```{r}
#| echo: false
#| eval: true
library(tidyverse)
library(tidyverse)
dat %>% 
  ggplot(aes(x = m, y = W)) + 
  geom_point() +
  geom_smooth(method = "lm",
              linetype = "dashed",
              color = "black",
              se = TRUE,
              linewidth = 1,
              alpha = 0.4)+
  theme_minimal()
```

## Endogeneity
- Remedies : 
  - theoretical grounding : adding predictors (like the weighing machine calibration)
  - We will see other methods that ensure exogeneity later

## Heteroskedasticity {.scrollable}
- Now imagine that your machine gives good estimates on average, but is much less reliable for higher masses
- The variance of the error changes with X
- $\sigma^2$ is not constant
- This would be heteroskedastic
```{r}
#| echo: false
#| eval: true
library(tidyverse)
library(tidyverse)

m <- sample(1:100, 100, replace = FALSE) / 10
W <- 9.81 * m + rnorm(100,0, 2*m)

dat <- data.frame(m, W)
lm_model <- lm(W ~ m, data = dat)
summary(lm_model)
```

## Heteroskedasticity
```{r}
#| echo: false
#| eval: true
library(tidyverse)
library(tidyverse)

m <- sample(1:100, 100, replace = FALSE) / 10
W <- 9.81 * m + rnorm(100,0, 2*m)

dat <- data.frame(m, W)
lm_model_biased <- lm(W ~ m, data = dat)

dat %>% 
  ggplot(aes(x = m, y = W)) + 
  geom_point() +
  geom_smooth(method = "lm",
              linetype = "dashed",
              color = "black",
              se = TRUE,
              linewidth = 1,
              alpha = 0.4)+
  theme_minimal()
```

## Heteroskedasticity {.scrollable}
- It can change the values of your standard errors
- Hence your hypotheses: run "robust" linear models instead
```{r}
#| echo: true
#| eval: false
library(performance)
library(MASS)
plot(model$fitted.values, model$residuals)
check_heteroscedasticity(model) # Run a Breush-Pagan test
robust_model <- MASS::rlm(y ~ x, data = y)

```
```{r}

library(performance)
library(MASS)
library(stargazer)

x <- sample(1:1000, 1000, replace = "TRUE")

y <- 10*x + rnorm(1000, 0, 10)
y <- data.frame(y, x)
model1 <- lm(y ~ x, data = y)

y <- 10*x + rnorm(1000, 0, 10*(x))
y <- data.frame(y, x)
model2 <- lm(y ~ x, data = y)

y %>% 
  ggplot(aes(x = x, y = y)) + geom_point()+
  geom_smooth(method = "lm", linetype = "dashed", color = "black", se = TRUE, linewidth = 1, alpha = 0.4)+
  theme_minimal()

mean(model2$residuals)
plot(model2$fitted.values, model2$residuals)
check_heteroscedasticity(model2) # Run a Breush-Pagan test

model2 <- lm(y ~ x, data = y)
# Fit robust regression
robust_model <- MASS::rlm(y ~ x, data = y)

stargazer(model1, model2, robust_model, align = TRUE, 
          type = "text")
```

## Multicollinearity {.scrollable}
- Two predictors are extremely close, or several predictors summed give another predictor
- E.g. in the case of the weights : 
  - you add volume in the predictors :

\begin{equation}
W_i = \gamma + \alpha m_i + \beta V_i + \epsilon_i
\end{equation}
- You can use `check_collinearity(model)`
- variance inflation factor (VIF) : VIF > 10 is highly colinear
```{r}
#| echo: true
library(performance)

set.seed(123)

m <- sample(1:100, 100, replace = FALSE) / 10

# density of copper ≈ 8920 kg/m^3
v <- m/8920 + rnorm(100, 0, 0.00001)  # small measurement error

y <- 9.81*m + rnorm(100, 0, 10)

model1 <- lm(y ~ m)
model2 <- lm(y ~ m + v)

summary(model1)
summary(model2)

check_collinearity(model2)
```

## Remedies
- You should theoretically discuss your choice of predictors!
  - here, you know that mass is the driver of weight, and not volume... hence remove volume

- In more tricky cases, you can use principal component analysis to determine your main predictor (not covered here)

## Another example
```{r}
#| echo: true
library(performance)
model1 <- lm(far_right_relvotes ~ pagri+pindp+pcadr+ppint+pempl+pouvr+sup, data = elections2024 %>% filter(dens == "Urbain intermédiaire"))
model2 <- lm(far_right_relvotes ~ pcadr+pouvr+dens, data = elections2024)

summary(model1)
summary(model2)

check_collinearity(model1)
check_collinearity(model2)

```
# Interactions in linear regressions

## What we saw so far {.scrollable}

\begin{equation}
Y_i = (\beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3 +...+\beta_k X_k + \epsilon)_i
\end{equation}

- Predictors can be categorical (t-test) or continuous
- what if they interact ?
- e.g. does being rich has the same effect on the vote for the far-right in lower-educated and higher-educated communes ?

\begin{equation}
Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \gamma X_1 X_2 + ... + \epsilon
\end{equation}

- Any increase $\delta X_1$ leads to $\delta X_1 (\beta_1 + \gamma X_2)$ increase in $Y$

## How to run and interpret {.scrollable}
- You just add `lm(Y ~ X_1+X_2 + X_1*X_2)`
- Use `ggpredict()` to visualise and interpret

```{r}
#| echo: true
#| eval: false

elections2024 <- elections2024 %>% filter(dens == "Urbain intermédiaire") %>% 
               mutate(petr = etranger/pop, psup = sup/pop, pnodip = nodip/pop, log_revmoyfoy = log(revmoyfoy))

model3 <- lm(data = elections2024,
             far_right_relvotes ~ log_revmoyfoy +psup + log_revmoyfoy*psup +petr+ pagri+pouvr+ pcadr+age+log(prixbien)+pempl+ppint+pindp+psup+pnodip)

library(ggeffects)

ggpredict(model3, terms = c("log_revmoyfoy", "psup")) |> 
  as_tibble() |> 
  mutate(
    group = factor(
      group,
      levels = c("0.14", "0.23", "0.32"),
      labels = c(
        "Low share of higher education (14%)",
        "Medium share of higher education (23%)",
        "High share of higher education (32%)"
      )
    )
  ) |> 
  ggplot(aes(x, predicted, color = group, fill = group)) + 
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  labs(
    color = "Share with higher education",
    fill  = "Share with higher education"
  )
```

## You get with continuous variables
```{r}
#| echo: false
#| eval: true

elections2024 <- elections2024 %>% filter(dens == "Urbain intermédiaire") %>% 
               mutate(petr = etranger/pop, psup = sup/pop, pnodip = nodip/pop, log_revmoyfoy = log(revmoyfoy))

model3 <- lm(data = elections2024,
             far_right_relvotes ~ log_revmoyfoy +psup + log_revmoyfoy*psup +petr+ pagri+pouvr+ pcadr+age+log(prixbien)+pempl+ppint+pindp+psup+pnodip)

library(ggeffects)

ggpredict(model3, terms = c("log_revmoyfoy", "psup")) |> 
  as_tibble() |> 
  mutate(
    group = factor(
      group,
      levels = c("0.14", "0.23", "0.32"),
      labels = c(
        "Low share of higher education (14%)",
        "Medium share of higher education (23%)",
        "High share of higher education (32%)"
      )
    )
  ) |> 
  ggplot(aes(x, predicted, color = group, fill = group)) + 
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  labs(
    color = "Share with higher education",
    fill  = "Share with higher education"
  )
```


# Non-linear models
## Motivation
- Linear regressions are good for continuous DVs
- What if Y is binary and your IV continuous ?
- Many dependent variables are binary in policy evaluation...
- school enrollment, hospital admissions, etc. can be coded as 0 and 1s at the individual level

## Your scatter plot
```{r}
# Load necessary library
# Load necessary library
library(ggplot2)

set.seed(123)

# Simulate continuous independent variable
n <- 300
x <- rnorm(n, mean = 0, sd = 1)

# Create probability using logistic function
true_prob <- 1 / (1 + exp(-( -1 + 2*x )))

# Generate binary outcome
y <- rbinom(n, size = 1, prob = true_prob)

data <- data.frame(x = x, y = y)

# Fit models
linear_model  <- lm(y ~ x, data = data)
logit_model   <- glm(y ~ x, data = data, family = binomial)

# Create prediction grid
x_seq <- seq(min(x), max(x), length.out = 200)
pred_data <- data.frame(x = x_seq)

pred_data$linear_pred <- predict(linear_model, newdata = pred_data)
pred_data$logit_pred  <- predict(logit_model, newdata = pred_data, type = "response")

# Plot
ggplot(data, aes(x = x, y = y)) +
  geom_jitter(height = 0.05, alpha = 0.4) +
  labs(title = "Linear vs Logistic Regression for Binary Outcome",
       subtitle = " ",
       x = "Continuous Independent Variable",
       y = "Binary Outcome (0/1) - (e.g. voting)") +
  theme_minimal()

```

## A linear regression 
```{r}
# Load necessary library
# Load necessary library
library(ggplot2)

set.seed(123)

# Simulate continuous independent variable
n <- 300
x <- rnorm(n, mean = 0, sd = 1)

# Create probability using logistic function
true_prob <- 1 / (1 + exp(-( -1 + 2*x )))

# Generate binary outcome
y <- rbinom(n, size = 1, prob = true_prob)

data <- data.frame(x = x, y = y)

# Fit models
linear_model  <- lm(y ~ x, data = data)
logit_model   <- glm(y ~ x, data = data, family = binomial)

# Create prediction grid
x_seq <- seq(min(x), max(x), length.out = 200)
pred_data <- data.frame(x = x_seq)

pred_data$linear_pred <- predict(linear_model, newdata = pred_data)
pred_data$logit_pred  <- predict(logit_model, newdata = pred_data, type = "response")

# Plot
ggplot(data, aes(x = x, y = y)) +
  geom_jitter(height = 0.05, alpha = 0.4) +
  geom_line(data = pred_data, aes(y = linear_pred), 
            color = "red", size = 1.2) +
  labs(title = "Linear vs Logistic Regression for Binary Outcome",
       subtitle = "Red = Linear Regression | Blue = Logistic Regression",
       x = "Continuous Independent Variable",
       y = "Binary Outcome (0/1)") +
  theme_minimal()

```

## A linear regression
- Violates several assumptions about linear models
- Heteroskedastic
- Irrelevant

## But there are probabilities... {.scrollable}
- Take different intervals of X
```{r}
library(ggplot2)

set.seed(123)

# Simulate continuous independent variable
n <- 300
x <- rnorm(n, mean = 0, sd = 1)

# Create probability using logistic function
true_prob <- 1 / (1 + exp(-( -1 + 2*x )))

# Generate binary outcome
y <- rbinom(n, size = 1, prob = true_prob)

data <- data.frame(x = x, y = y)

# ---- Create 3 equal-width intervals of x ----
breaks <- seq(min(data$x), max(data$x), length.out = 4)
data$x_group <- cut(data$x, breaks = breaks, include.lowest = TRUE)

# Fit models
linear_model  <- lm(y ~ x, data = data)
logit_model   <- glm(y ~ x, data = data, family = binomial)

# Create prediction grid
x_seq <- seq(min(x), max(x), length.out = 200)
pred_data <- data.frame(x = x_seq)

pred_data$linear_pred <- predict(linear_model, newdata = pred_data)
pred_data$logit_pred  <- predict(logit_model, newdata = pred_data, type = "response")

# Plot
ggplot(data, aes(x = x, y = y, color = x_group)) +
  geom_jitter(height = 0.05, alpha = 0.6) +
  geom_line(data = pred_data, aes(y = logit_pred), 
            color = "blue", size = 1.2) +
  labs(title = "Linear vs Logistic Regression for Binary Outcome",
       subtitle = "Red = Linear Regression | Blue = Logistic Regression",
       x = "Continuous Independent Variable",
       y = "Binary Outcome (0/1)",
       color = "X Interval") +
  theme_minimal()

```

- it is continuous :

```{r}
ggplot(data, aes(x = x, y = y, color = x)) +
  
  # Points
  geom_jitter(height = 0.06, size = 2.5, alpha = 0.8) +
  
  # Logistic regression (main focus)
  geom_line(data = pred_data, 
            aes(x = x, y = logit_pred),
            inherit.aes = FALSE,
            color = "#2C3E50",
            linewidth = 1.6) +
  
  # Linear regression (lighter, secondary)
  geom_line(data = pred_data, 
            aes(x = x, y = linear_pred),
            inherit.aes = FALSE,
            color = "#E74C3C",
            linewidth = 1.2,
            linetype = "dashed") +
  
  # Beautiful continuous gradient
  scale_color_viridis_c(option = "plasma", direction = -1) +
  
  labs(
    title = "Why Logistic Regression Fits Binary Data Better",
    subtitle = "Solid curve = Logistic model | Dashed line = Linear model",
    x = "Continuous Predictor (x)",
    y = "Probability / Binary Outcome",
    color = "Value of x"
  ) +
  
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "right"
  )
```
## A logistic regression
```{r}
# Load necessary library
# Load necessary library
library(ggplot2)

set.seed(123)

# Simulate continuous independent variable
n <- 300
x <- rnorm(n, mean = 0, sd = 1)

# Create probability using logistic function
true_prob <- 1 / (1 + exp(-( -1 + 2*x )))

# Generate binary outcome
y <- rbinom(n, size = 1, prob = true_prob)

data <- data.frame(x = x, y = y)

# Fit models
linear_model  <- lm(y ~ x, data = data)
logit_model   <- glm(y ~ x, data = data, family = binomial)

# Create prediction grid
x_seq <- seq(min(x), max(x), length.out = 200)
pred_data <- data.frame(x = x_seq)

pred_data$linear_pred <- predict(linear_model, newdata = pred_data)
pred_data$logit_pred  <- predict(logit_model, newdata = pred_data, type = "response")

# Plot
ggplot(data, aes(x = x, y = y)) +
  geom_jitter(height = 0.05, alpha = 0.4) +
  geom_line(data = pred_data, aes(y = linear_pred), 
            color = "red", size = 1.2) +
  geom_line(data = pred_data, aes(y = logit_pred), 
            color = "blue", size = 1.2) +
  labs(title = "Linear vs Logistic Regression for Binary Outcome",
       subtitle = "Red = Linear Regression | Blue = Logistic Regression",
       x = "Continuous Independent Variable",
       y = "Binary Outcome (0/1)") +
  theme_minimal()

```

## A logistic regression
```{r}
# Load necessary library
# Load necessary library
library(ggplot2)

set.seed(123)

# Simulate continuous independent variable
n <- 300
x <- rnorm(n, mean = 0, sd = 1)

# Create probability using logistic function
true_prob <- 1 / (1 + exp(-( -3 + 10*x )))

# Generate binary outcome
y <- rbinom(n, size = 1, prob = true_prob)

data <- data.frame(x = x, y = y)

# Fit models
linear_model  <- lm(y ~ x, data = data)
logit_model   <- glm(y ~ x, data = data, family = binomial)

# Create prediction grid
x_seq <- seq(min(x), max(x), length.out = 200)
pred_data <- data.frame(x = x_seq)

pred_data$linear_pred <- predict(linear_model, newdata = pred_data)
pred_data$logit_pred  <- predict(logit_model, newdata = pred_data, type = "response")

# Plot
ggplot(data, aes(x = x, y = y)) +
  geom_jitter(height = 0.05, alpha = 0.4) +
  geom_line(data = pred_data, aes(y = linear_pred), 
            color = "red", size = 1.2) +
  geom_line(data = pred_data, aes(y = logit_pred), 
            color = "blue", size = 1.2) +
  labs(title = "Linear vs Logistic Regression for Binary Outcome",
       subtitle = "Red = Linear Regression | Blue = Logistic Regression",
       x = "Continuous Independent Variable",
       y = "Binary Outcome (0/1)") +
  theme_minimal()

```

## Logistic regression
- Instead of a linear function 
\begin{equation}
Y = \beta_0 + \beta_1 X
\end{equation}
- We assume that Y has a predicted probability that is a non-linear function of X :
\begin{equation}
\mathcal{P}(Y = 1) = f_\mathcal{L}(\beta_0 + \beta_1 X) = \frac{1}{1 + e^{(-(\beta_0 + \beta_1 * X))}}
\end{equation}
- The goal is to find $\beta_0, \beta_1$ !

## Different logistic regressions
```{r}
library(ggplot2)

# Logistic function
logistic <- function(x, b0, b1) {
  1 / (1 + exp(-(b0 + b1 * x)))
}

# Create x grid
x <- seq(-5, 5, length.out = 400)

# Define parameter combinations
params <- data.frame(
  b0 = c(0, 0, -2, 2, 1),
  b1 = c(1, 2, 1, 1, -2),
  label = c("β0=0, β1=1 (baseline)",
            "β0=0, β1=2 (steeper)",
            "β0=-2, β1=1 (shift right)",
            "β0=2, β1=1 (shift left)",
            "β0=1, β1=-2 (negative slope)")
)

# Generate predictions
plot_data <- do.call(rbind, lapply(1:nrow(params), function(i) {
  data.frame(
    x = x,
    prob = logistic(x, params$b0[i], params$b1[i]),
    label = params$label[i]
  )
}))

# Plot
ggplot(plot_data, aes(x = x, y = prob, color = label)) +
  geom_line(size = 1.2) +
  labs(title = "Effect of β0 and β1 in Logistic Regression",
       x = "X",
       y = "Predicted Probability",
       color = "Parameter Values") +
  theme_minimal() +
  ylim(0,1)

```

## Finding these parameters
- Done through so-called maximum likelihood estimation, based on your data ! 
- $\beta_0$, $\beta_1$ is for one predictor
- You can have k predictors :
\begin{equation}
\mathcal{P}(Y = 1) = \frac{1}{1 + e^{(-(\beta_0 + \beta_1 * X_1+\beta_2 X_2 + \beta_3 X_3 +...+ \beta_k X_k))}}
\end{equation}
- that can be continuous or categorical !


## Logistic regression in R {.scrollable}
- You can perform a logistic regression in R using `glm`:
```{r}
#| eval: false
#| echo: true

logit_model <- glm(binary_var ~ continuous_var, 
                   data = your_data,
                   family = 'binomial')
summary(logit_model)
```

```{r}
#| eval: true
#| echo: false
set.seed(42)

n <- 400

# Nicely scaled continuous predictor
continuous_var <- rnorm(n, mean = 0, sd = 1)

# Moderate true relationship
# Intercept = -0.5
# Slope = 1.2  (reasonable effect size)
true_prob <- plogis(-0.5 + 1.2 * continuous_var)

# Binary outcome
binary_var <- rbinom(n, size = 1, prob = true_prob)

data <- data.frame(
  continuous_var = continuous_var,
  binary_var = binary_var
)
```

```{r}
#| eval: true
#| echo: false
library(tidyverse)
logit_model   <- glm(binary_var ~ continuous_var, data = data, family = binomial)
summary(logit_model)
```

- You need a binary DV ! 
- How can you interpret ?


## Coming back to our made-up data... {.scrollable}
```{r}
#| echo: true
#| eval: true
library(ggeffects)

# like the predcit() function of Base R, we use ggpredict() and specify
# our variable of interest
predicted_probs <- ggpredict(logit_model, terms = "continuous_var")
predicted_probs
```

```{r} 
#| echo: true
#| eval: false
# this is the simplest way of plotting this
ggplot(predicted_probs, aes(x = x, y = predicted)) +
  # our graph is more or less a line, so geom_line() applies
  geom_line(color = "blue", size = 1) +
  geom_point() +
  # geom_ribbon() with the values that ggpredict() provided for the confidence
  # intervals then gives
  # us a shade around the geom_()line as CIs
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), fill = "grey", alpha = 0.3) +
  labs(
    title = "Predicted Probability of Voting by Political Interest",
    x = "Political Interest",
    y = "Predicted Probability of Voting"
  ) + 
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )
```

## Predicted probabilities

```{r} 
#| echo: false
#| eval: true
# this is the simplest way of plotting this
ggplot(predicted_probs, aes(x = x, y = predicted)) +
  # our graph is more or less a line, so geom_line() applies
  geom_line(color = "blue", size = 1) +
  geom_point() +
  # geom_ribbon() with the values that ggpredict() provided for the confidence
  # intervals then gives
  # us a shade around the geom_()line as CIs
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), fill = "grey", alpha = 0.3) +
  labs(
    title = "Predicted Probability",
    x = "Continuous variable",
    y = "Predicted probability"
  ) + 
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )
```