---
title: Session 1 — Descriptive statistics
toc: true
toc-depth: 3
editor: 
  markdown: 
    wrap: 72
---

## Exercice 1 - manual warm up 1
  Imagine having 3 dices (one is blue, the other green and the last one is red) with 3 faces each, and values going from 1 to 3. Consider the sum of the three dices as random variable $X_{bgr}$:
  
1. Compute the probability of obtaining 5, with the blue dice being 2, the red being 2, and the green being 1: $P(X_{bgr} = 5_{212})$
2. What is the set of potential outcomes, $S_X$ of $X_{bgr}$ ?
  3. Now imagine that the dices can not be distinguished anymore, your random variable is $X$. What is your probability of obtaining 3, $P(X=3)$ ?
  4. What is the probability of obtaining 5 ?
  5. How much is the expectation ?
  6. You know that one of your dices is worth 3. What is the probability of obtaining 5 in total?
  7. Draw a bar plot, showing the potential outcomes in the x axis, and the probabilities in the y axis.
  8. What is the conditional expectation worth in the case where one of the three dices is worth 3 ? Is that equal to the non-conditional expectation of X ?
  9. Verify all your results by simulating the rolling of these three dices, 10000 times, in R.


## Exercice 2 - Playing with RStudio
  Open RStudio on your personal computer (or your personal coding environment).
  
  1. Create a new project named "Session 1 - Methods R&PE", in a new directory. What are the different files created ? Quit the session, go to your folder on your personal computer, and click on the '.Rproj' file created. What happens ?
  2. Write 1+1 in your main .R script and press "Run".
  3. Create a new folder in this project named 'data'.
  4. Install packages "tidyverse" and "rio" in the console.
  5. You're all set for importing data and start creating functions/the next exercices ! 


## Exercice 3 - playing with data
The [European social survey](https://www.europeansocialsurvey.org/about-ess) "measures the attitudes, beliefs and behaviour patterns of diverse populations in more than thirty nations". Use the `rio` package to import data from <"https://jan-rovny.squarespace.com/s/ESS_FR.dta">. 

  1. Use `glimpse` function to have a look at all. Restrict your data set to `euftf` (views on European integration), `gndr` (gender), `happy`, `yrbrn`(birthyear) and `eduyrs` (years of education), using function `select()` from `tidyverse`.
  2. How many observations are there ? Are the variables categorical or continuous ?  
  3. Summarise your data set and comment on the findings, using `table()` and  `summary()`. Drop the NA values. 
  4. Create a new variable `age`, with the age of the respondents, using `mutate()`. Use 2026 as reference year.
  5. Plot the distribution of the variables `euftf` and `happy` using `barplot(table())`. What can you say about these distributions? What is the most common answer ? What is the percentage of respondents who have given this answer ?
  6. Plot a histogram of the age of the respondents, with 10 breaks. What is the decade with the most respondents ? 
  7. Plot the distribution of `happy` for `gndr == 1` and `gndr == 2`, using `ggplot2 : ggplot()` and ` geom_bar(position = "dodge")`
  8. Does being male/female influence your chances of self-reporting being happy ? Conduct a `t-test` to test the hypothesis that gender influences people's self assessed level of happiness. What are the 95% and 99% confidence intervals ?


```{r}
#| echo: true
#| eval: true
ESS<-import("https://jan-rovny.squarespace.com/s/ESS_FR.dta")
glimpse(ESS)
ESS <- ESS %>% select(euftf, happy,yrbrn, eduyrs, gndr)
length(ESS$euftf)
summary(ESS)
ESS <- ESS %>% mutate(age = 2026-yrbrn)
barplot(table(ESS$euftf)/length(ESS$euftf))
hist(ESS$age, breaks = 10)
ggplot(ESS, aes(x = happy, fill = factor(gndr))) +
    geom_bar(position = "dodge") + theme_minimal()

t_test_result <- t.test(
  ESS %>% filter(gndr == 1),
  ESS %>% filter(gndr == 2),
  alternative = "two.sided",
  conf.level = 0.99,
  var.equal = FALSE  # Use Welch's t-test for unequal variances
)
t_test_result
```
## Exercice 4 : the median income in French rural communes
You want to analyse the electoral outcomes of the radical right in rural communes in France. For this, you have data on the 2024 European elections at <https://github.com/paulservais-1/methods_policyevaluation/raw/main/slides/session1/data/elections2024.csv>, and data on the French municipalities at <https://github.com/paulservais-1/methods_policyevaluation/raw/main/slides/session1/data/demographics2024.xlsx>. 

  1. Use `rio` to import both data sets, stored in two different variables
  2. Use `glimpse` to . Look at the types of codedep and codecommune. Can you merge on `codedep` and `codecommune` in these conditions ? Use `as.numeric()` for both to prepare for the merge
  3. Use the `left_join()` function to merge both data set on codecommune and codedep.
  4. Restrict your data frame to rural communes, using `filter`.
  5. Replace NA values by the mean of the variable.
  6. What is the range of the far-right vote ? in absolute and relative terms ? 
  7. What is the average of the relative far-right vote in French rural communes ? The median ? The standard deviation ?
  8. Plot the distributions of far-right voting in the different kinds of French rural communes (present in `dens7`), using `ggplot()`.
  9. Is the expected far-right voting significantly different in those different rural typologies ? Compute different t-tests, and conclude.



```{r}
#| echo: true
#| eval: true

library(rio)
elec_eur <- import("https://github.com/paulservais-1/methods_policyevaluation/raw/main/slides/session1/data/elections2024.csv")
socio_dem <- import("https://github.com/paulservais-1/methods_policyevaluation/raw/main/slides/session1/data/demographics2024.xlsx")

glimpse(elec_eur)
glimpse(socio_dem)

socio_dem$codecommune<- as.numeric(socio_dem$codecommune)
socio_dem$codedep <- as.numeric(socio_dem$codedep)

elec_eur <- elec_eur %>% left_join(socio_dem)
elec_eur <- elec_eur %>% filter(dens == "Rural")

for(variable in colnames(elec_eur)){
  if(is.numeric(elec_eur[[variable]])) {
    elec_eur[[variable]] <- ifelse(is.na(elec_eur[[variable]]),
                                   mean(elec_eur[[variable]], na.rm = TRUE),
                                   elec_eur[[variable]])
  }
}

summary(elec_eur)

ggplot(data = elec_eur, aes(x = far_right_relvotes, color = dens7))+geom_density()

# Example numeric variable (replace with your actual variable)
numeric_var <- elec_eur$far_right_relvotes

# Split the data by dens7 categories
group1 <- numeric_var[elec_eur$dens7 == "Rural à habitat dispersé"]
group2 <- numeric_var[elec_eur$dens7 == "Bourgs ruraux"]
group3 <- numeric_var[elec_eur$dens7 == "Rural à habitat très dispersé"]

# Perform pairwise t-tests
t_test_1_2 <- t.test(group1, group2)
t_test_1_3 <- t.test(group1, group3)
t_test_2_3 <- t.test(group2, group3)

# Print results
print(t_test_1_2)
print(t_test_1_3)
print(t_test_2_3)


```

## Take-home assignement
  Find any data set online related to your research interests, or originating from the replication package of a paper that you would like to replicate. 
  
  1. Select two variables and describe them using `summary`, and `` (or their Stata/Python equivalent)
  
  2. Plot their distribution
  
  3. Use a binary variable splitting your variable in two groups, and conduct a t-test to compare the means of those two groups. Comment on your findings.