---
title: Session 5 - Regression discontinuity design
toc-depth: 3
editor: 
  markdown: 
    wrap: 72
---

## Load the relevant packages
Start by loading the following packages:
```{r}
#| eval: false
#| echo: true

library(rio)
library(tidyverse)
library(rdrobust)
library(rddensity)
```

## Exercice 1: regression discontinuity design (RDD)

This exercise is inspired by Carpenter and Dobkin (2015). You are provided with a data set about crime rates in the US, which you can find at <https://github.com/scpo-quantimethods/quantitative_methods_two/raw/main/exercices/exercices5/data/carpenter_dobkin_data.dta>. You want to analyse the effect of alcohol on crime rates.

1. Import the data. The codebook is the following :
- `age` stands for the age
- `arrested_for` qualifies the reasons for which people have been arrested. `DUI` stands for driving under the influence.
- `number` is the number of people of that age arrested for that reason.

2. Run the following code : 

```{r}
#| eval: false
p <- data_set %>%  ggplot(aes(x = age, y = number)) +
  geom_point(alpha = 0.05) +
  facet_wrap(facets = vars(arrested_for), scales = "free_y") +
  theme_minimal()

p
```

3. Do you observe any “jump” or discontinuity in your graphs? If so, where?

4. Restrict your dataset to DUIs (driving under the influence).

5. Run your regression discontinuity using rdrobust, with age as the independent variable and number as the dependent variable, using a cutoff at 21 and a bandwidth of 2.

6. Is this a sharp or fuzzy RDD?

7. Does being allowed to drink alcohol affect the number of DUIs?

8. Repeat the analysis using the number of assaults and the legal majority at 18 to examine whether turning 18 has an impact on the number of assaults.

9. Run a McCrary density test to verify that there is no sorting at the cutoff. Verify that your results are robust for different polynomial degrees.
```{r}
#| eval: false
#| echo: false

library(rio)
library(tidyverse)
library(rdrobust)
library(estimatr)
library(AER)

carpenter_dobkin_prepared <- import("data/carpenter_dobkin_data.dta")
glimpse(carpenter_dobkin_prepared)
p <- carpenter_dobkin_prepared %>%  ggplot(aes(x = age, y = number)) +
  geom_point(alpha = 0.05) +
  facet_wrap(facets = vars(arrested_for), scales = "free_y") +
  theme_minimal()

p

carpenter_dobkin_dui_only <-
  carpenter_dobkin_prepared |>
  filter(
    arrested_for == "DUI",
    abs(age - 21) < 2
  )

carpenter_dobkin_assault_only <-
  carpenter_dobkin_prepared |>
  filter(
    arrested_for == "Assault",
    abs(age - 18) < 2
  )

rdd_dui <- rdrobust(
  # specifies the DV
  y = carpenter_dobkin_dui_only$number,
  # specifies the running variable of the RDD
  x = carpenter_dobkin_dui_only$age,
  # specifies the cutoff
  c = 21,
  # specifies the bandwidth, here the unit of analysis is years
  h = 2,
  all = TRUE
) 

rdd_assault <- rdrobust(
  # specifies the DV
  y = carpenter_dobkin_assault_only$number,
  # specifies the running variable of the RDD
  x = carpenter_dobkin_assault_only$age,
  # specifies the cutoff
  c = 18,
  # specifies the bandwidth, here the unit of analysis is years
  h = 2,
  all = TRUE
) 

rdd_dui |> summary()
rdd_assault %>% summary()
```


## Exercice 2

This exercise explores the causal effect of narrowly winning an election on the candidate’s performance in the next election. The idea is that winning by just over 50% allows the candidate to run as an incumbent in the following election, yet candidates having very close to the 50% (below or above) are in practice very comparable in terms of ressources. We can therefore compare candidates just above 50%, to candidates who lost by just under 50% to identify the incumbency advantage.

You are provided with a dataset data_set (simulated for this exercise), which you can find here <https://github.com/paulservais-1/methods_policyevaluation/raw/main/data/incumbency.dta>. Each row corresponds to a candidate, and each column to its electoral results in election t, and in the next election t+1. The codebook is :

- `vote_share_t` : vote share (%) for the candidate in the first election
- `vote_share_t1` : vote share (%) for the same candidate in the following election

1. Import the data set

2. Create a variable `incumbent` worth 1 when the candidate is an incumbent in the second election

3. Plot the relationship between vote_share_t and vote_share_t1, and change the color at the cutoff.

4. Do you observe a jump anywhere ? If yes, where is it located?

5. Estimate a regression discontinuity to estimate the incumbency effect on electoral outcomes.

- Using function `lm` and restricting your data to different bandwidths.
- Using rdrobust().
How large is the effect ? How significant is the effect ?

6. Is it a sharp or fuzzy RDD? 

7. Repeat the RDD estimation with different polynomial orders (p = 1, 2, 3). Are the results sensitive to the polynomial order?

8. Run a McCrary density test to check for potential sorting around the cutoff. Can we assume that candidates do not manipulate their vote share to just cross 50%?


## Exercice 3

We want to study whether entering a PhD program in Grandes Ecoles increases individuals’ future earnings. Admission is largely determined by the grade on the master’s thesis:

- Candidates with a thesis grade above 16/20 are more likely to be admitted.
- Some candidates below 16 are admitted occasionally, and some above 16 are rejected.

You are provided with a data set following individuals over 10 years, containing :
1. `thesis_grade`: their score at the master's thesis
2. Their `future_income`, ten years later.
3. Their admission status in the PhD program, admitted.

It can be found here <https://github.com/paulservais-1/methods_policyevaluation/raw/main/data/grades_ge.dta>.

1. Import the data set.
2. Run the following code :

```{r}
#| eval: false
library(ggplot2)
library(dplyr)
data_set <- data_set %>% mutate(admit_jitter = admitted + runif(n, -0.05, 0.05))
ggplot() +
  geom_point(data = data_set, aes(x = thesis_grade, y = admit_jitter), alpha = 0.4) +
  labs(
    title = "Fuzzy RDD: Probability of PhD Admission vs. Thesis Grade",
    x = "Thesis Grade",
    y = "Admission status"
  ) +
  theme_minimal()
```
Do you notice any effect of the grade on the admissions ? At which cutoff ?

3. Plot the relationship between income and the running variable, and change the color at the cutoff.

4. Do you observe a jump anywhere ? If yes, where is it located?

5. Estimate a regression discontinuity to estimate the effect of being admitted in a Grande Ecole on future earnings.

- Using function `lm` and restricting your data to different bandwidths.
- Using rdrobust().
How large is the effect ? How significant is the effect ?

6. Is it a sharp or fuzzy RDD? 

7. Repeat the RDD estimation with different polynomial orders (p = 1, 2, 3). Are the results sensitive to the polynomial order?

8. Run a McCrary density test to check for potential sorting around the cutoff. Can we assume that candidates do not manipulate their master's thesis grade just above the cutoff?