---
title: Session 1 — Descriptive statistics
toc: true
toc-depth: 3
editor: 
  markdown: 
    wrap: 72
---

# Introduction

The core of any research in social science is to discover and describe
relationships between variables. Whether you describe a research puzzle
as a "why X despite Y?" type of question, or as finding the
relationships between independent and dependent variables, it generally
involves unveiling causation relationships previously hidden, unknown or
ignored by analysts. The social facts Y that the researchers seek to
explain are generally called the *dependent variable*, the
*explanandum*, while the variables supposed to explain it are called the
*independent variables* or the *explanans*.

Of course, many methods exist in practice to make such claims about
connection between variables. These methods are generally divided in two
large sub-groups : qualitative methods, which are generally based on
small-N comparisons and can for instance allow to better access
interpretive dimensions, and quantitative methods. The latter methods
allow to conduct large N comparisons, to replicate analyses, and have
known very important recent developments related to technological
improvements in the recent decades (e.g. Computational Social Science,
LLMs). Similarly, these methods generally seek to find a. That is, a
given variable $Y$, is caused (and can eventually be predicted) by a
series of other variables $X_1, X_2, X_3, X_4$, etc. Hence, in a very
general form, considering that $Y$ can be explained by
$X_1, X_2, X_3, X_4$ means that we assume that there exists a function
$f : \mathbb{R}^4 \rightarrow \mathbb{R}$ that satisfies : $$
\begin{equation}
Y = f(X_1,X_2, X_3, X_4)  + \epsilon 
\end{equation}
$$

Hence, any research process typically involves the following stages :
collection of data on variables $Y$, $X_1, X_2, X_3, X_4$, analysis of
these data, and drawing interpretations and conclusions based on these
data. The data are generally classified in two subgroups: categorical
and continuous (or interval) variables. Categorical variables have a
limited amount of potential values, and for instance refer to
**nominal** measures which help qualitatively describe the observations,
with no logical ordering - e.g. the gender of an interviewee can be male
= 1, female = 2, else = 3, etc. - or **ordinal** measures, with a
logical ordering - e.g. a variable that qualifies the level of
religiosity of someone on a 0 to 5 scale.

As such, as being the first step, the **data collection phase is crucial
in every empirical research project** - which is generally the case in
political science or sociology. This data collection phase typically
follows the following steps :

-   **Download the data**, typically from online sources : national
    institutes of statistics (e.g. INSEE in France, in the UK, in the
    USA) , official ministries - -, or collection - eg. through survey
    experiments, web scraping, or any other method -). This step also
    comes with selecting the desired format of the data (typically .csv,
    .xslx, .dat, .spss, .stata, .Rdata, or parquet), as well as reading
    the meta data, which allow to understand where the data is from,
    when it was done, for what purpose, etc.
-   **Import the data** : this involves importing data in your
    environment to be able to manipulate it.
-   **Inspect data** : this involves looking at your data in your
    environment, analyse the table
-   **Data cleaning** : this step generally comes with treating the NAs,
    remove clear outliers. In practice, most data set are provided in a
    rather coarse fashion, and the researcher must transform it to make
    it suitable for his/her analysis.
-   **Descriptive statistics** : this step comes with analysing the
    important dimensions that can help describe quantitatively your
    data: the means, the standard deviation, the median, the mode, the
    interquartile ranges, the maximum and minimums, and make bar plots,
    density plots or histograms, to analyse the dispersion or skewness
    of the data.

Ideally, the data collected should be both reliable and valid : that
means, it should have a small random error (the errors affecting the
measurements themselves), and limited systematic error (your measure
accurately measures the concept you seek to analyse).

Take for instance happiness as a variable to explain : as a social
scientist, you would like to understand what causes people to be happy
or not. Of course, you might have first struggles in defining
happiness - an essentially subjective measure -, and you might as well
have competing hypotheses about what causes happiness. In addition, you
might have random errors perturbating your data collection process.

## Download data

Once you have a research question that you seek to analyse, you may want
to download data relevant for your purposes. Imagine that I want to
analyse the impact of religiosity on happiness. This would require me to
find a data set that has both information on :

1.  Individual people

2.  Their level of religiosity - your independent variable (you seek to
    analyse if religiosity impacts happiness)

3.  Their level of happiness - your dependent variable (you seek to
    explain happiness)

4.  Ideally, other variables characterising the respondent's profile :
    his/her gender, his/her age, his/her income, etc., which will help
    isolate the specific effect of religiosity on happiness.

Finding the data set is not a trivial operation : it requires prior
knowledge on data sets available out theser, remaining alert to new
issues, and a lot of walking around on the internet to find data that
could be relevant to your research problem. Discussions with peers are
very useful for that. After discussion with friends or your research
supervisor, you might for instance learn that such lab has issued such
data set, this ministry provides such information, or this statistics
institute provides these data. Your

## Import the data in your local environment

You may thus now have your data stored locally in your computer, in what
case you should provide the path for your environment (RStudio or Stata)
to access it. Otherwise, your computer will not know what you are
talking about and probably return you a "No such file or directory" type
of error. A good practice is therefore to store the data that your are
using in a subfolder called "data", located in the same folder as where
your piece of code or .RProject lies.

```{r}
library(tidyverse)
# Importing data using the absolute path : gives the actual location of the data on your computer


# Importing data using the relative path : gives the location of the data on your computer, relative to your code script

# Importing data from the internet 

```

Of course, the function used to read your data will depend on the format
of your data itself. Here are a couple other examples for other formats
than .csv.

```{r}
# .xslx (Excel)


# .dat

# .stata

```

## Inspect the data

Once your data is opened in your local environment, you can start
inspecting it. A good start is to inspect the first and last rows :

```{r}
head_data <- head(data_ess, 15) #refer the data set, refer the number of rows to keep
tail_data <- tail(data_ess, 15) #refers the last rows
head_data
tail_data

```

This shows that the data is structured in columns and rows : each row
corresponds to one observation/measurement, and each column corresponds
to a variable.

You may want to know the names of the variables and have a specific look
at some of the interesting variables for your purposes

```{r}
colnames(data_ess)
head(data_ess %>% select(), 15)
unique(data_ess %>% select()) # shows the different values contained in your column
table(data_ess %>% select())  # shows the different values and their count
```

# Basics of probability theory

## Probability, distribution functions

Imagine that you have two dices with six faces on each dice. Your random
variable is the outcome of each throw of dices.

You would like to assess your chances of obtaining each value contained
in the set of potential outcomes $S = \{2,3,4,5,6,7,8,9,10,11,12\}$.
Intuitively, you might want to start by looking at the 2 value : 2 is
only obtained in the case where you obtain one for the two dices. That
is, for dice 1, you have $1/6$ of obtaining the value, and then again
$1/6$ of obtaining 1 for the second dice. Overall, the probability of
obtaining $2$, $P(Y=2)$ is therefore : $$
\begin{equation}
P_2 = 1/6 \times 1/6 = 1/36
\end{equation}
$$ Of course, if you now want to look at the probability of obtaining
$3$ as outcome, $P( Y = 3)$ your probability will be the union of two
probabilities : the chance of obtaining 1 with dice 1 and 2 with dice 2,
or the one of obtaining 2 with dice 1 and 1 with dice 2, which are
clearly equivalent because the dices cannot be distinguished.
Mathematically, this writes : $$
\begin{equation}
P(Y=3) = 1/6 \times 1/6 + 1/6 \times 1/6 = 2/36 = 1/18
\end{equation}
$$ You see that the probability is doubled, compared to P(Y=2).

If you look at the joint probability, this corresponds to a diagonal :
2. Hence, you see that obtaining 7 is the last diagonal and therefore $$
\begin{equation}
P(Y=7) = 6\times1/36 = 1/6
\end{equation}
$$

\begin{array}{c|cccccc}
 & 1 & 2 & 3 & 4 & 5 & 6 \\
\hline
1 & \frac{1}{36} & \frac{1}{36} & \frac{1}{36} & \frac{1}{36} & \frac{1}{36} & \frac{1}{36} \\
2 & \frac{1}{36} & \frac{1}{36} & \frac{1}{36} & \frac{1}{36} & \frac{1}{36} & \frac{1}{36} \\
3 & \frac{1}{36} & \frac{1}{36} & \frac{1}{36} & \frac{1}{36} & \frac{1}{36} & \frac{1}{36} \\
4 & \frac{1}{36} & \frac{1}{36} & \frac{1}{36} & \frac{1}{36} & \frac{1}{36} & \frac{1}{36} \\
5 & \frac{1}{36} & \frac{1}{36} & \frac{1}{36} & \frac{1}{36} & \frac{1}{36} & \frac{1}{36} \\
6 & \frac{1}{36} & \frac{1}{36} & \frac{1}{36} & \frac{1}{36} & \frac{1}{36} & \frac{1}{36}
\end{array}

Imagine now that you are skeptical of these theoretical findings, and
you decide to carry a small experiment in order to test this. Let's
imagine that you are feeling very bored for an entire year and that you
decide to launch these two dices a hundred thousand times, and report
results in two tables $X_1, X_2$ with the result for each dice. Your
results might look like :

```{r}
library(DT)
N = 100000 #number of measurements
x1 = sample(1:6,N, replace=T) # first dice (values are from 1 to 6)
x2 = sample(1:6,N, replace=T) # second dice (values are from 1 to 6)
head(x1)
```

In fact your final result - the sum of the two dices -, is the union of
two independent events $X_1, X_2$ being respectively the outcome of dice
$1$ and $2$, provided that they are indeed independent. Your final
result $Y$ is therefore simply the sum of these two random variables.
You can then sum :

```{r}
y = x1+x2 # total number obtained
freq <- table(y)
barplot(freq)
```

and the function `table()` gives you the number of occurences for each
value. The probability to obtain 7 from your sample is then simply the
number of occurences that you have measured, divided by the total number
of attempts that you have done, in this case 10000. Hence :

```{r}
prob <- freq/N
prob
barplot(prob)
```

These values are called the probability distribution of the variable Y.
That is, for each outcome of Y, there is a probability to obtain it
(here shown on the barplot). These probabilities follow the simple rule
: $$
\begin{equation}
\sum_{i=1}^N P(Y = Y_i) = P(Y = Y_1)+P(Y= Y_2) +... P(Y = Y_n) = 1 
\end{equation}
$$ That is, the sum of all probabilities over the set of potential
outcomes $\{Y_1,Y_2,Y_3,..., Y_n\}$ is equal to 1. The $\Sigma$ symbol
simply stands for sum. Now imagine that you want to know the average
value $$
\begin{equation}
\mathbb{E}(Y) = \sum_{i=1}^N Y_i \ * P(Y_i) = \bar{Y}
\end{equation}
$$

This value is called the expectation.

## Conditional probability, conditional expectation

Imagine now that, due to the extreme fatigue provoked by the repetitive
throwing of those dices, you throw those a bit too wildly and one falls
under the table where you are. The other one is there and shows the
value 3. You want to know the expected value of $y$, knowing that one
dice among the two is 3. This is called a conditional probability and
noted $P(y|X_1)$, with $| X_1$ simply meaning that $X_1$ is considered
fixed (here worth 3). Since $X_1$ is fixed, the probability is
$P(y|X_1) = 1/6$, and knowing this information is actually very precious
on Y, because it considerably reduces the set of potential outcomes : $$
Y \ge 4 \text{, } Y\le 9 \text{ and } S_Y = \{4, 5, 6, 7, 8, 9\} 
$$ $$
\begin{equation}
\mathbb{E}(Y|X_1 = 3) = \sum_{i=1}^N Y_i\ * P(Y_i|X_1 = 3) 
\end{equation}
$$ and so we get : $$
\begin{align}
  \mathbb{E}(Y|X_1 = 3) = 1/6 \times (4 + 5 + 6 + 7 + 8 + 9) = 6.5\\
  \mathbb{E}(Y|X_1 = 4) = 1/6 \times (5 + 5 + 7 + 8 + 9 + 10) = 7.5
\end{align}
$$ Notice that the expected value of y therefore changes, as a function
of the explanatory variable, $X_1$.

## Ceteris paribus hypothesis and bias

This conditional expectation is central in quantitative social science,
as it is generally what the researcher seeks to evaluate, when
confronted to social facts. Coming back to the example of the evaluation
of happiness, imagine that you are similarly capable of measuring the
level of happiness (self-reported on a 1 to 5 scale) of 10000
individuals. If you are interested to know whether happiness is
influenced by the degree of religiosity of an individual, you might want
to ask them questions about their practicing of religion $X$, and then
to know how $Y$ evolves with $X$, on average. This would lead you to be
interested in the expectation :\
$$
\begin{equation}
\mathbb{E}(Y) = f(X)
\end{equation}
$$ This information would already be of an interesting value, yet it
would not allow you to isolate the specific causation relation between
happiness, and religiosity, as it is very likely the case that
religiosity is perhaps correlated with something else, like income, that
might both explain happiness and religiosity. As such, the expectation
that does not account for that might underestimate or overestimate the
role of religiosity on happiness. That is why you would rather be
interested in knowing whether an individual's religiosity impacts its
happiness level, all else being equal - which is sometimes called the
*ceteris paribus* hypothesis (latin words for all else being equal),
central in econometrics. That is, in this last case, if income is the
only variable that could introduce spuriousness or bias between
happiness and religiosity, you might want to know : $$
\begin{equation}
\mathbb{E}(Y|X_1) = f(X_2|X_1)
\end{equation}
$$ Which basically means that you are interested in how variable Y is
affected by $X_2$, for a given variable $X_1$ - hence, this isolates the
effect of $X_2$, from the potential confounding variable $X_1$.

## Towards continuity

Now simply imagine that you repeat that process with dices that have
$n = 1000, 1400, 800$ faces, with each value ranging from $1/n$ to $ 1$.
And that you would like to know what are the final probabilities of
having any decimal number between 0 and 3. Of course, the manual
computation of this would be very cumbersome - which is why automatising
the process with R can prove very useful. Reusing the same code would
give :

```{r}
library(ggplot2)
library(dplyr)
N = 100000 #number of measurements
x1 = sample(1:1000,N, replace=T)/1000 # first dice (values are from 1 to 6)
x2 = sample(1:1400,N, replace = T)/1400
x3 = sample(1:800,N,replace = T)/800
y = x1+x2+x3
dat <- data.frame(y)
head(dat, 10)
ggplot(dat, aes(x = y)) +
  geom_histogram(aes(y = after_stat(density))) +
  geom_density(color = "red", linewidth = 1) +
  labs(title = "Distribution of y = x1 + x2 + x3",
       x = "y",
       y = "Density") +
  theme_minimal()
```
Here, you clearly see that your variable is now almost continuous, and your probability to obtain one specific value becomes extremely small. We therefore generally focus on the probability of having $Y\le Y_0$, that is, the sum of probabilities of obtaining all the values below 4 (included). Coming back to our example of dices : for one dice, your probability of obtaining a result $\le 4$ is just the sum of the probabilities of getting all the values before 4 : 
\begin{equation}
P(X_1 \le 4) = 1/6 + 1/6 + 1/6 + 1/6 = 4/6
\end{equation}
For a continuous variable, this sum is noted as an integral - yet it means the exact same thing  : 
\begin{equation}
P(Y \le Y_0) = \int_\underline{R_Y}^Y_0 f_Y(y') \text{d}y'
\end{equation}

# Real examples
## Real-life examples

Of course in practice dices are not extremely interesting... and you may want to analyse real data, describing actual social facts. Yet the same logic will apply to real data ! Imagine for instance working on the French communes: you seek to describe the median income in those communes, but you only have access to a random sample of 10.000. You will need a coding environment : RStudio (open source, free, huge community) and data... : your task will be to find it online.


## Coding environment
The R language is perhaps : R (CRAN). Its advantage over other softwares, like SPSS, is that it is entirely free and open-source, and that is maintained by a very large community of users. A very nice environment where to code on R is RStudio - which provides different panels where to organise your code, create new projects, etc. If you need help to download [R](https://cran.r-project.org/) and [RStudio](https://posit.co/downloads/), please check Malo Jan's webpage <https://malo-jn.quarto.pub/introduction-to-r/session1/0102_setup.html>, or other sources like <https://rstudio-education.github.io/hopr/starting.html>. In general, the R documentation, StackOverflow and AI are you friends in these cases is very useful !
  
## Creating a project 
The first step when you launch RStudio is to create a project in which you can work. The best way to do that is to go in `File > New Project > New directory > New Project > Create a new directory`. In the last step, you can choose the location of your directory. This will create a folder, in which your project will be contained. There, you will be able to create new scripts (".R" files), where you will be able to code. You can also code directly in the console. If you need reminders on the basic usage of R, please visit <https://malo-jn.quarto.pub/introduction-to-r/>.

## Data and packages
R comes with basic language and functions : base R. This basic infrastructure for instance allows you to add, multiply, conduct basic operations in your files. In general, you want to perform many more operations, which is why I strongly recommend the use of packages and libraries. These packages are already coded functions and tools that allow you to be much more efficient in your code. Among all those libraries, `tidyverse` is **extremely** useful. You can load and install such packages in this way :
```{r}
#| echo: true
#| eval: false
install.packages("tidyverse")
library(tidyverse)
```

## Importing data

Other useful packages are those that allow to download and import data. Imagine that you have data retrieved from the internet : here a file containing information on the french communes - demographics2024.xlsx. This file is an "excel" file, hence it should be read by a program capable to read excel formats: this requires the readxl package. Another more general function is `import()`, which comes with the package `rio`. This allows you to disregard the format of your file (it is coded to support various formats). You can also export() data using rio.
```{r}
#| echo: true
#| eval: false
library(readxl) # needed to read excel without "rio"
communes <- read_excel("data/demographics2024.xlsx") # relative path
communes <- read_excel("/Users/Desktop/Website/methods_policyevaluation/slides/session1/data/") # absolute path
#using the "rio" package
library(rio)
communes <- import("data/demographics2024.xlsx")
```


```{r}
#| echo: false
#| eval: true

library(readxl)
library(tidyverse)
communes <- read_excel("data/demographics2024.xlsx") # relative path

len_com <- length(communes %>% pull(codecommune))
sample_com <- sample(1:len_com, N, replace = FALSE)
comms_rand <- NULL
a <- 0
for(i in sample_com){
  a =a+1
  comms_rand[a] = (communes %>% pull(codecommune))[i]}
comm_r <- communes %>% filter(codecommune %in% comms_rand) 
```

Once you have data imported, you can rapidly inspect it, by using the function `head()` and `tail()`, which allow you to get the first and last rows of a data set, see the different variables it contains and their type. 
```{r}
#| echo: true
head(comm_r, 3)
tail(comm_r, 3)
length(comm_r$codecommune)
```
This shows you for instance that you have communes going from departement 01 to 95, for the Year 2024, with different socio-demographic information about the occupational structure, the median income, etc. 

## Categorical variable
For categorical variables, you can check their different values in the following way:
```{r}
#| echo: true
# Nominal : the density typology of communes 
unique(comm_r$dens)
table(comm_r$dens)
summary(comm_r$dens)
```
Here, we show the number of communes from the sample that are 'rural', 'urbain intermédiaire' or "urbain dense'. `unique` gives you the different values present in your variable, while `table` returns you the different values present in the data set and their number of occurences. Note that `summary()`, which is extremely useful for continuous variables, here returns the type of your variable, because it is non-numeric.

## Interval variable
If you look at a continuous variable - e.g. the median income in communes -, you can use the function `summary()` to obtain essential informations on the variable : its mean, median, its quartiles, and the number of NA's. NA's are gaps in your variable, missing information - that are therefore coded as Not Assessed. You may either remove these, by using `drop_na()`, or fill these, by adding zeros instead or the mean of the variable. This can be done in this way :
```{r}
#| echo: true
# Continuous : the median income in those communes
summary(comm_r$revmoyfoy)
range(comm_r$revmoyfoy, na.rm = TRUE)
summary((comm_r %>% drop_na(revmoyfoy))$revmoyfoy) #dropping NA's
#replacing the NA's
comm_r <- comm_r %>% mutate(revmoyfoy = ifelse(is.na(revmoyfoy), mean(revmoyfoy, na.rm =TRUE), revmoyfoy))
summary(comm_r$revmoyfoy)
```
Notice that your last summary no longer contains NA values, and has an unchanged mean value - which is the objective of filling with mean values instead of the NA's.

## Density
You can also easily have a look at the density, by using `ggplot2` package, that is extremely useful for plots and graphs in R. `ggplot` works this way: you just have to specify your data (here `comm_r`), your variables (typically x, y and color) in Aesthetic Mapping (aes) `aes(x =, y =, color = )` and then add different shapes to your graph : here `geom_density()`. You can also try plotting a histogram with `geom_histogram()` !

```{r}
#| echo: true
ggplot(comm_r, aes(x= revmoyfoy, color = dens)) + geom_density()
```

## Skewness 
Something we have not discussed so far and is quite important is the notion of skewness of a distribution. To introduce you to this notion, take the following example : you analyse the age distribution in a small school bus. The following ages are present in the bus :

\begin{equation}
\text{Ages} =\{13,15,15,16,16,17,13,15,17, 18, 55\},
\end{equation}

Where most ages correspond to pupils, and the last one is the bus driver. If you compute the median, it is worth 16 (the value that splits your distribution in two chunks with equal sizes), while if you compute the mean, $\bar{\text{Age}} = 19.09$. Hence, your mean is very sensitive to extreme values ! That is why your distribution can be 'skewed', and non-symmetrical, if your median is different than your mean. You typically find the three cases : 
- Median $<$ Mean : positively skewed (the mean is pushed upwards by high values)
- Median $>$ Mean : negatively skewed (the mean is pushed downwards by low values)
- Mean = Median : symmetrical distribution

In the previous case, for instance, the distributions are positively skewed : you see in the distribution that some very high values are pushing the mean upwards, compared with the median. In order to bring down these extreme values, you can use transformations, like logarithms, to make a positively skewed variable look more symmetrical:
```{r}
#| echo: true
ggplot(comm_r, aes(x= log(revmoyfoy), color = dens)) + geom_density()
```

## Expectations
R also allows you to very easily compute expectations. In the case of median income : 
\begin{equation}
\text{E}(\text{MI}) = \frac{1}{10000}\sum_{i=1}^{10000} \text{MI}_i, 
\end{equation}
is simply this way: 
```{r}
#| echo: true
mean(comm_r$revmoyfoy)
```
And the conditional expectations can also simply be computed by filtering out the values that are undesired.
```{r}
#| echo: true
mean((comm_r %>% filter(dens == "Rural"))$revmoyfoy)
mean((comm_r %>% filter(dens == "Urbain dense"|dens == "Urbain intermédiaire"))$revmoyfoy)
```
Where you clearly see that :

\begin{equation}
\text{E}(\text{MI}| \text{dens} = \text{Rural}) \le \text{E}(\text{MI}| \text{dens} = \text{Urbain}) 
\end{equation}

Should we conclude that median income is significantly smaller in rural communes than in urban areas ? The density plot indeed shows that rural areas are more concentrated in the lower income regions, compared with urban areas, but the dispersion being quite large for urban areas... it is hard to know whether this difference is statistically significant or not ! The typical way to compare these two means is to use a t-test. 

## Student's t-test
Imagine that I would like to verify, lets say with a confidence margin of 99%, that 
\begin{equation}
\text{E}(\text{MI}| \text{dens} = \text{Rural}) \le \text{E}(\text{MI}| \text{dens} = \text{Urban}) 
\end{equation}
The t-test consists in computing a t-statistic comparing the means of the two independent samples : 
\begin{equation}
t = \frac{\bar{X_1} - \bar{X_2}}{\sqrt{\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}}}
\end{equation}
Under the assumption that the two variables have normal distributions. If the two means are indeed statistically similar, that means that the t value should be non discernible from 0, with the 99% level of confidence decided. In the case where they are different, the t-value should be significantly different than 0. That means, the confidence interval defined at the 99% confidence level : 
\begin{equation}
(\bar{X_1} - \bar{X_2}) \pm t_{99%} * \sqrt{\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}}
\end{equation}
should not include 0. 
In R, you can easily implement that using :

```{r}
rural_income <- comm_r %>% filter(dens == "Rural") %>% pull(revmoyfoy)
urban_income <- comm_r %>% filter(dens == "Urbain dense" | dens == "Urbain intermédiaire") %>% pull(revmoyfoy)

t_test_result <- t.test(
  rural_income,
  urban_income,
  alternative = "two.sided",
  conf.level = 0.99,
  var.equal = FALSE  # Use Welch's t-test for unequal variances
)
```
Where you test the `NULL` hypothesis, which is that the means are similar - hence that the difference is not statistically different than 0. If your p-value is below 0.01, you conclude that your hypothesis is rejected.
The result of the t-test ran above indeed shows you that your 99% confidence interval [-7135.543; -5418.664] excludes 0, which means that the test is indeed significant, and you can clearly reject the NULL hypothesis with 99% confidence (even much more) : the means are different, and the average difference is 6277, 11 euros (36 640,58 - 30363,47).

